---
format:
  html:
    code-summary: "Show the code"
---

# Clustering of Spatial Transcriptomics Data

```{r}
#| label: setup
#| include: false

knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)

# set the working directory to the course_files folder
# make sure to run course_files/download_course_files.sh
knitr::opts_knit$set(root.dir = here::here("course_files"))
```

::: {.callout-tip}
#### Learning Objectives

- TODO
:::

## Setup

We continue working with the sagittal mouse brain dataset from previous chapters.

::: {.callout-note collapse="true"}
#### Click to expand

Start by loading the required libraries and the Seurat object.

```{r}
#| label: load-visium-data

# Load libraries
library(Seurat) # single-cell and spatial analysis toolkit
library(sparseMatrixStats)
library(paletteer) # colour palettes
library(ggplot2) # plotting
library(dplyr) # data manipulation
library(patchwork) # combining plots

# Load the Seurat object from the previous chapter
visium <- readRDS("precomputed/mouse_brain_clustering.rds")
```

As a reminder, this object was created by:

- Importing the raw data using `Load10X_Spatial()`, followed by QC and filtering to remove low-quality spots.
- Data normalisation using `SCTransform()`, which models technical variation and applies variance stabilisation.
  Stored in the "SCT" assay.
- Applying dimensionality reduction using PCA, UMAP and t-SNE.
- Clustering the data using graph-based clustering.
  The clusters we use in this chapter were generated using the Leiden algorithm at a resolution of 0.8 and were set the the default cell identities (`Idents(visium)`).
:::

## Overview

Following from our clustering, a natural next step is to understand what each cluster represents biologically.
To do this, we identify **marker genes**, i.e. genes that are differentially expressed between the clusters.

It is worth noting that this procedure is statistically flawed and can lead to misleading conclusions.
The reason being that our clusters were defined based on gene expression patterns, so we are not testing independent hypotheses.
In other words, we should expect to find significant markers for each cluster, as the clusters themselves were defined based on gene expression pattern differences.
This is true, even if the clusters are not biologically meaningful.
For that reason, this analysis should be considered **exploratory and hypothesis-generating**, rather than a rigorous statistical test.
The marker genes we identify can be used to assign biological identity to the clusters, but this should be done with caution and ideally **validated using external data or experiments**.

There are several statistical tests used for marker gene identification.
For the reason mentioned above, we shouldn't worry too much about being "statistically rigorous" here, so simple approaches are often sufficient.
In fact, one of the most commonly used tests is the **Wilcoxon rank-sum test**, which is a generic non-parametric test that compares the ranking of expression of each gene in one cluster against the others.
This test is robust to outliers and does not assume a specific distribution.

Tests specifically designed for single-cell data (e.g. MAST) can better account for variable detection rates across cells, but can be computationally more intensive.
For spatial transcriptomics data with fewer cells and often lower sparsity than single-cell data, the Wilcoxon test typically performs well and is sufficiently fast.

## Find Marker Genes

Seurat provides the function `FindAllMarkers()` to perform differential expression analysis between a target cluster and all other clusters.

Some critical parameters of this function to consider are:

- **`only.pos`**: If `TRUE`, only returns markers that are more highly expressed in the cluster of interest.
- **`min.pct`**: The minimum percentage of cells/spots in the cluster where the gene is detected (default 0.1).
  Raising this threshold prevents genes expressed in just a few spots from dominating the marker list.
- **`logfc.threshold`**: the minimum log-fold change in expression for a gene to be considered for testing.

Both `min.pct` and `logfc.threshold` are should be set to relatively permissive values, as their main purpose is to reduce the computational burden of testing genes that are unlikely to be markers.
These parameters trade off sensitivity (finding all true markers) against specificity (avoiding false positives).
Stricter thresholds produce fewer, higher-confidence markers, while relaxed thresholds result in more markers but may include noisy results.

Let's apply `FindAllMarkers()` to our Visium data and explore the results.

```{r}
#| label: find-markers

# Identify marker genes for each cluster
# Expressed in at least 5% of spots in the cluster, with a log-fold change of at least 0.25
markers <- FindAllMarkers(
  visium,
  only.pos = TRUE,
  min.pct = 0.25,
  logfc.threshold = 0.25
)

# The result is a standard data.frame
head(markers)
```

The result of this function is a standard R data frame with the following columns:

- `pval` is the p-value from the statistical test
- `avg_log2FC` is the average log2 fold change of the gene in the cluster compared to others
- `pct.1` is the percentage of spots in the cluster where the gene is detected
- `pct.2` is the percentage of spots in the other clusters where the gene is detected
- `p_val_adj` is the p-value adjusted for multiple testing
- `cluster` is the cluster for which the gene is a marker
- `gene` is the name of the gene

## Filtering Results

We can manipulate the output table using `dplyr` (part of `tidyverse`), to combine several criteria to find the most potentially interesting markers.

For example, in the code below, we filter for significant markers (adjusted p-value < 0.05).
Then, for each cluster, we take the top 5 genes with the highest log-fold change.

```{r}
#| label: top-markers

# View top 5 markers for each cluster
top5 <- markers |>
  # filter for significant markers (adjusted p-value < 0.05)
  filter(p_val_adj < 0.05) |>
  # then for each cluster
  group_by(cluster) |>
  # take the top 5 genes with the highest log-fold change
  slice_max(avg_log2FC, n = 5) |>
  # sort them by cluster and descending log-fold change
  arrange(cluster, desc(avg_log2FC)) |>
  # reset grouping
  ungroup()

# Genes in cluster 1
top5 |>
  filter(cluster == 1) |>
  pull(gene)
```

We can visualise the expression of these marker genes across the tissue using `SpatialFeaturePlot()`, which shows the spatial distribution of gene expression.

```{r}
#| label: spatial-feature-top5

# Visualise the 5 top markers for cluster 1
feature_plot <- SpatialFeaturePlot(
  visium,
  features = top5 |> filter(cluster == 1) |> pull(gene),
  ncol = 2
)

# Visualise the spatial location of cluster 1
cluster_plot <- SpatialDimPlot(
  visium,
  cells.highlight = WhichCells(visium, idents = "1")
) +
  ggtitle("Cluster 1 Spatial Location")

# Plot side by side
feature_plot + cluster_plot
```

Another way you can filter this table is based on the `pct.1` and `pct.2` columns, which represent the fraction of spots where the gene was detected in the target cluster and in other clusters, respectively.
For example, below we set the very strict criteria that the gene has to be detected in all spots of the target spot, and in no more than 50% of the rest of the spots:

```{r}
#| label: filter-by-pct

# Filter on percent detected
top_by_pct <- markers |>
  filter(p_val_adj < 0.05 & pct.1 == 1 & pct.2 < 0.5)

# Count how many genes we have per cluster
top_by_pct |>
  count(cluster)
```

This only returns `r nrow(top_by_pct)` in only a few of the clusters.
We could visualise these to see how specifically localised they are in our tissue:

```{r}
#| label: spatial-feature-top-pct

# Visualise the top markers based on pct filtering
SpatialFeaturePlot(
  visium,
  features = top_by_pct$gene,
  ncol = 3
)
```

We could follow-up by comparing these markers to literature or reference datasets, to start assigning biological meaning identity to these clusters.

## Summary

::: {.callout-tip}
#### Key Points

- TODO
:::
