# Deconvolution

::: {.callout-caution}
This section is under revision
:::

```{r}
#| label: setup
#| include: false

knitr::opts_chunk$set(
  echo = TRUE,
  message = FALSE,
  warning = FALSE,
  cache.lazy = FALSE
)

# set the working directory to the course_files folder
# make sure to run course_files/download_course_files.sh
knitr::opts_knit$set(root.dir = here::here("course_files"))
```

::: {.callout-tip}
#### Learning Objectives

- Understand the concept of deconvolution in spatial transcriptomics
- Apply deconvolution methods to spatial transcriptomics data
- Use the `RCTD` package for deconvolution analysis
- Visualize and interpret deconvolution results
:::

## Setup

We continue working with the sagittal mouse brain dataset from previous chapters.

::: {.callout-note collapse="true"}
#### Click to expand

Start by loading the required libraries and the Seurat object.

```{r}
#| label: load-visium-data

# Load libraries
library(Seurat) # single-cell and spatial analysis toolkit
library(sparseMatrixStats)
library(paletteer) # colour palettes
library(ggplot2) # plotting
library(dplyr) # data manipulation
library(patchwork) # combining plots

# Load the Seurat object from the previous chapter
visium <- readRDS("precomputed/mouse_brain_banksy.rds")
```

As a reminder, this object was created by:

- Importing the raw data using `Load10X_Spatial()`, followed by QC and filtering to remove low-quality spots.
- Data normalisation using `SCTransform()`, which models technical variation and applies variance stabilisation.
  Stored in the "SCT" assay.
- Applying dimensionality reduction using PCA, UMAP and t-SNE.
- Clustering the data using graph-based clustering.
  The clusters we use in this chapter were generated using the Leiden algorithm at a resolution of 0.8 and were set the the default cell identities (`Idents(visium)`).
- Identifying spatially variable features using Moran's I and markvariogram methods.
- Running the BANKSY algorithm to identify spatial domains in the data, which were stored in the `banksy_clusters` column of the metadata.
:::

## Overview

Deconvolution is a computational technique used to estimate the cellular composition of bulk gene expression data.
In the context of spatial transcriptomics, deconvolution can help identify the proportions of different cell types within pixels/spots.
This is relevant, especially for low resolution technologies, where each spot likely covers multiple cells (e.g. Visium V1/V2 with spots of 55 μm diameter).
If those cells are of different types, then the gene expression profile of that spot will be a mixture of the profiles of the different cell types, which can confound downstream analysis such as clustering and tissue architecture analysis.

There are many computational tools for deconvolution of spatial transcriptomics data, reviewed in [Gaspard-Boulinc et al. 2022](https://doi.org/10.1038/s41467-022-29913-1).
Here, we will exemplify deconvolution using the Robust Cell Type Decomposion (RCTD) method of [Cable et al. 2022](https://doi.org/10.1038/s41587-021-00830-w), which was one of the top-performing methods in that benchmark study.

Briefly, RCTD works by using an annotated reference dataset (e.g. from single-cell data) and the query dataset (the spatial data to deconvolute).
The method then uses a statistical method to infer cell mixtures in the spatial data, based on the reference dataset.

RCTD is implemented in the [`spacexr`](https://www.bioconductor.org/packages/release/bioc/html/spacexr.html) package.
We load that package, along with the `pheatmap`, which we will use for heatmap visualisations.

```{r}
#| label: load-libraries

# Load necessary libraries
library(spacexr)
library(pheatmap)
```

## Reference Dataset

The first step for deconvolution is to find a suitable reference dataset.
The choice of reference dataset will depend on the tissue and organism you are working with, and may not always be available for non-model organisms.

In this case, we will use a preprocessed single-cell RNA-seq dataset from the [Allen Brain Cell (ABC) Atlas](https://brain-map.org/bkp/explore/abc-atlas).
These cells are annotated in a hierarchical manner: class → subclass → cluster, from broad cell categories to more specific cell populations.

- **Class**: Broad cell type category representing major neuronal and non-neuronal groups (e.g. glutamatergic neurons, GABAergic neurons, astrocytes, oligodendrocytes).

- **Subclass**: A division within a class, grouping cells with similar transcriptional profiles and biological characteristics (e.g. Vip, Sst, Lamp5 subclasses within GABAergic neurons).

- **Cluster**: The most specific annotation level, representing related cell populations identified from single-cell transcriptomic data.
  Cluster names are often based on marker genes that distinguish the population (e.g. *Vip Arhgap36 Hmcn1*).

Let's load the object and have a quick look at the data to understand its structure.
As this is a precomputed dataset, we will load it from an RDS file and then use the `UpdateSeuratObject()` function to ensure compatibility with the latest Seurat version.

```{r}
#| label: load-reference

# Load the reference single-cell dataset
ref <- readRDS("precomputed/mouse_brain_reference_for_RCTD.rds")
ref <- UpdateSeuratObject(ref)

# Assays present in object
Assays(ref)
```

As this is a single-cell dataset, it contains the "RNA" assay, which is the default assay containing the raw counts.
The data was also normalised using sctransform, which is stored in the "SCT" assay (set as the default).

The metadata contains three columns with the annotation levels: `class`, `subclass` and `cluster`.

```{r}
#| label: reference-annotations

# Check the annotation levels in the reference dataset
ref[[]] |>
  select(class, subclass, cluster) |>
  head()

# Count how many of each annotation level we have
length(unique(ref$class))
length(unique(ref$subclass))
length(unique(ref$cluster))
```

We can visualise these annotations on the UMAP to see how they relate to the clusters and to each other.

```{r}
#| label: reference-umaps
#| code-fold: true
#| code-summary: "Click to expand visualisation code"

# Visualise UMAP and Spatial plots of the reference dataset with different annotation levels
umap_class <- DimPlot(ref, reduction = "umap", group.by = "class", label = TRUE)
umap_subclass <- DimPlot(
  ref,
  reduction = "umap",
  group.by = "subclass",
  label = TRUE
)
umap_cluster <- DimPlot(
  ref,
  reduction = "umap",
  group.by = "cluster",
  label = FALSE
)

(umap_class + umap_subclass) /
  umap_cluster &
  theme(legend.position = "none") &
  coord_equal()
```

Based on this, we can decide at what level we want to run the deconvolution analysis.

For our analysis, we will use the subclass annotation for deconvolution (`r length(unique(ref$subclass))` categories), as it provides a good balance between granularity and interpretability.

You can count how many cells we have of each subclass in the reference dataset, as this will impact the deconvolution results.
If you use a different single-cell dataset, **make sure that each cell type has a sufficient number of cells** (at least 25) to ensure RCTD works reliably.

```{r}
#| label: reference-cell-counts
#| results: "hide"

# Count number of reference cells in each subclass
table(ref$subclass)
```

## Preparing RCTD Data

RCTD uses its own specific format for the reference and query data.

### Reference Object

The reference object requires counts, cluster identities, and number of UMIs for each cell to be extracted and formatted into a Reference object.

We extract these from our Seurat object and create the Reference object for RCTD:

```{r}
#| label: prepare-reference

# raw counts from the RNA assay
ref_counts <- ref[["RNA"]]$counts

# our cluster identities will be the subclass annotation
ref_cluster <- as.factor(ref$subclass)
names(ref_cluster) <- colnames(ref)

# the total number of UMIs for each cell
ref_umis <- ref$nCount_RNA
names(ref_umis) <- colnames(ref)

# Create the Reference object for RCTD
reference <- Reference(ref_counts, ref_cluster, ref_umis)

# Remove intermediate objects to free memory
rm(ref, ref_counts, ref_cluster, ref_umis)
```

### Query Object

Simiarly, we need to prepare the spatial transcriptomics data that we want to deconvolute with RCTD.
We will convert the Seurat object to a format compatible with RCTD, which requires:

- The raw counts from the "Spatial" assay
- The spatial coordinates of the spots
- The total number of UMIs for each spot

```{r}
#| label: prepare-query

# raw counts
query_counts <- visium[["Spatial"]]$counts

# spot coordinates
query_coords <- GetTissueCoordinates(visium)[, c("x", "y")]

# total UMIs for each spot
query_umis <- colSums(query_counts)

# Construct the query object for RCTD
query <- SpatialRNA(query_coords, query_counts, query_umis)

# remove intermediate objects to free memory
rm(query_counts, query_coords, query_umis)
```

## Running RCTD

Once we have our reference and query objects prepared, we can run the deconvolution analysis using RCTD.
This involves two steps:

- `create.RCTD()`: creates the RCTD object using the reference and query data.
  This step preprocesses and prepares the data for deconvolution.
  It calculates the gene expression profiles for each cell type in the reference dataset, filters genes to retain only those that are differentially expressed between the cell labels used for deconvolution, and estimates a normalisation factor to account for platform differences (e.g. single-cell versus single-nucleus).
- `run.RCTD()`: runs the deconvolution analysis, which estimates the proportions of different cell types in each spot based on the reference single-cell dataset and the spatial transcriptomics data.

Let's run the first step to create the RCTD object:

```{r}
#| label: create-rctd

# Create RCTD object on 8 cores
rctd <- create.RCTD(query, reference, max_cores = 8)

# Remove intermediate objects to free memory
rm(query, reference)
```

We used **`max_cores = 8`** to speed up the computation, which can be time-consuming for larger datasets.

Once the RCTD object is created, we can run the deconvolution analysis using the `run.RCTD()` function.
This function will estimate the proportions of different cell types in each spot based on the reference single-cell dataset and the spatial transcriptomics data.

RCTD has three modes for deconvolution, specified with the `doublet_mode` parameter:

- `"doublet"`: limits the cell type assignment to 1-2 types.
  This is usually suitable for higher resolution spatial technologies.
- `"multi"`: like doublet mode, but allows for more than 2 cell types, up to a maximum defined by the `MAX_MULTI_TYPES` parameter in the `create.RCTD()` function (4 by default).
- `"full"`: does not limit the number of cell types assigned to each spot.

The reason to include `doublet` and `multi` modes is that the model can sometimes overfit the data and assign more cell types to each spot than there are (e.g. with very small proportions).
Therefore, restricting the upper limit of cell types for each spot can help to get more reliable results.
The choice comes down to your knowledge of the tissue and the resolution of the techonology you are using.

For now, we will use the function in "doublet" mode, but later we look at the results in "full" mode.
Note that RCTD is a computationally demanding analysis.
For that reason, we provide a pre-processed file that you can read, but you can expand the code block below to see the code used for running it.

```{r}
#| label: run-rctd-doublet
#| cache: true
#| code-fold: true
#| code-summary: "Click for `run.RCTD()` command - takes a long time to run!"

# This will run about ~12-15 minutes on 8 cores
rctd <- run.RCTD(rctd, doublet_mode = "doublet")
```

```{r}
#| label: save-rctd-doublet
#| include: false

saveRDS(rctd, "precomputed/mouse_brain_rctd_doublet.rds")
```

```{r}
#| label: read-precomputed-rctd-doublet

# Read precomputed rctd object in doublet mode
rctd <- readRDS("precomputed/mouse_brain_rctd_doublet.rds")
```

This concludes the deconvolution analysis with RCTD.
We can now add the results back into the Seurat visium object and visualise them.
Because we ran RCTD in doublet mode, we have two sets of results: one for singlets and one for doublets.
We will check how many spots are doublets and compare both results.

## Extracting deconvolution results

The RCTD object contains several components from the analysis, which we will extract to a data frame and then add to the metadata of our Seurat object.
The key components we will extract at this stage are:

- The cell type assignment to each spot
- The *weights* associated with each type, which are effectively the estimated proportions of each cell type

These are stored within the `rctd@results` slot of the object, and we extract both into a data frame:

```{r}
#| label: rctd-doublet-results-df

# Main results data frame - retain only some columns of interest
doublet_results <- rctd@results$results_df |>
  select(spot_class, first_type, second_type)

# Add the associated with each type
doublet_results <- cbind(doublet_results, rctd@results$weights_doublet)

# Ensure column names are unique
names(doublet_results) <- c(
  "spot_class",
  "first_type",
  "second_type",
  "first_weight",
  "second_weight"
)

# Inspect the table
head(doublet_results)
```

We have the following columns:

- `spot_class` indicates the type of spot classification: "singlet", "doublet_uncertain", "doublet_certain" and "reject".
- `first_type` and `second_type` indicate the two types of cells identified in the spot (using the categories from the `subclass` variable we used in the reference dataset).
- `first_weight` and `second_weight` indicate the proportion of `first_type` and `second_type` cells in the spot, respectively.

We can start by counting how many cells we have of each class:

```{r}
#| label: count-doublet-class

# Count spot classes
doublet_results |>
  count(spot_class)
```

We have no spots classified as "singlet", which is expected given the spot size of Visium v1 and the complexity of the tissue.
Most spots are classified as "doublet_certain", which means that the model is confident that there are at least two cell types in those spots.
A smaller proportion of spots are classified as "doublet_uncertain", which means that the model is less confident about the presence of two cell types in those spots.
Finally, a small proportion of spots are classified as "reject", which means that the model was not able to assign a cell type to those spots with confidence.

We can add this to the metadata of our Seurat object for some visualisations.
For example, we can visualise where each spot is classified as a doublet or not.

```{r}
#| label: add-doublet-to-seurat

# Add results back into visium object
visium <- AddMetaData(visium, metadata = doublet_results)

# Check where each spot is classified as a doublet
# We use a custom colour scheme to better distinguish the different classes
SpatialDimPlot(visium, group.by = "spot_class") +
  scale_fill_manual(
    values = c(
      "reject" = "red",
      "doublet_certain" = "steelblue",
      "doublet_uncertain" = "orange"
    )
  ) +
  ggtitle("Spot classification by RCTD")
```

We can also visualise the first and second cell type identified by RCTD, and compare it to the clusters we identified earlier using Leiden.

```{r}
#| label: visualise-doublet-types
#| code-fold: true
#| code-summary: "Click for visualisation code"

# Spatial plot for first type
first_type_plot <- SpatialDimPlot(
  visium,
  group.by = "first_type"
) +
  ggtitle("First cell type identified by RCTD")

# Spatial plot for second type
second_type_plot <- SpatialDimPlot(
  visium,
  group.by = "second_type"
) +
  ggtitle("Second cell type identified by RCTD")

# Leiden clusters
leiden_cluster_plot <- SpatialDimPlot(
  visium,
  group.by = "Leiden_08",
  label = TRUE,
  label.size = 4
) +
  theme(legend.position = "none") +
  ggtitle("Clusters (Leiden)")

# Assemble the plots together
(first_type_plot + second_type_plot) / leiden_cluster_plot
```

## Full cell type analysis

We have so far only looked at the major and secondary cell type identified by RCTD in doublet mode.
Doublet mode is useful to identify the two most likely cell types in each spot and best suited for datasets where most spots contain only one or two cell types.

This is likely not the case for the Visium v1 data we are using here, as each spot likely contains multiple cell types based on the spot size of 55 μm diameter.

However, RCTD provides an additional **"full"** mode which deconvolves each spot to any number of cell types.
Although it might sound like a good idea to always run the model in "full" mode, it is important to remember that the model may overfit the data, so it's not a bad idea to restrict to doublets when working with high-resolution data.
In our case, however, it makes sense to attempt running the analysis in full mode.

As this takes a long time to run, we provide a pre-computed file for you to read in.
But you can see the code used to produce it below.

```{r}
#| label: run-rctd-multi
#| cache: true
#| code-fold: true
#| code-summary: "Click for `run.RCTD()` command - takes a long time to run!"

# Run RCTD in full mode
rctd_full <- run.RCTD(rctd, doublet_mode = "full")
```

```{r}
#| label: save-rctd-multi
#| include: false

saveRDS(rctd_full, "precomputed/mouse_brain_rctd_full.rds")
```

```{r}
#| label: read-precomputed-rctd-multi
#| eval: false

# Read precomputed rctd object in doublet mode
rctd_full <- readRDS("precomputed/mouse_brain_rctd_full.rds")
```

### Extract and visualise cell weights

When running in full mode, the result is provided as a matrix of weights for all cell types in the reference dataset, which are stored in `rctd@results$weights`.
This is a matrix with the same number of rows as the number of spots and the same number of columns as the number of cell types in the reference dataset, where each value represents the proportion of that cell type in that spot.
This can be useful for a more full exploration of the results, or to focus the analysis on specific cell types of interest.

The code below shows you how to extract these results from the RCTD object, and store them in a data frame for plotting with ggplot2.
We plot the weights for all cell subclasses to get a full overview of the classification results from RCTD.

```{r}
#| label: visualise-weights

# Extract weights
rctd_weights <- rctd_full@results$weights

# Add coordinates
rctd_weights <- cbind(rctd_weights, GetTissueCoordinates(visium)[, c("x", "y")])

# Convert to tibble and pivot to long format
rctd_weights <- rctd_weights |>
  as_tibble(rownames = "spot_barcode") |>
  tidyr::pivot_longer(
    -c(spot_barcode, x, y),
    names_to = "subclass",
    values_to = "weight"
  )

# Visualise weights for each subclass
# Note we're capping the colour scale to 0.5 to better visualise differences
rctd_weights |>
  ggplot(aes(x, y, colour = weight)) +
  geom_point(size = 0.1) +
  facet_wrap(~subclass) +
  theme_void() +
  scale_colour_viridis_c(limits = c(0, 0.5))
```

As always, you should **interpret these results grounded in biological knowledge** of the tissue and cell types you are looking at.

For example, take cell type "L2_3_IT" (which stands for [layer 2/3 intratelencephalic neurons](https://knowledge.brain-map.org/celltypes/CCN202002013/CS202002013_183)):

```{r}
#| label: highlight-rctd-cells
#| code-fold: true

# Highlight a particular cell type
SpatialPlot(
  visium,
  cells.highlight = WhichCells(visium, expression = first_type == "L2_3_IT")
)
```

We can see these occur in the supperficial cortex layers, which is what we might expect for this cell type.
However, we also see them in inner tissues, which might not be expected.
This could be due to the fact that the model is overfitting the data and assigning this cell type to spots where it is not actually present.
It could also be that our reference dataset does not contain all the cell types present in the tissue, and therefore the model is assigning the closest cell type from the reference to those spots.

This is why it's important to **carefully interpret the results of deconvolution and use additional information to validate the results** (e.g. known marker genes, spatial location, etc.).

### Add cell weights to Seurat

As we did before with our doublet detection mode, we can also add these weights to the Seurat metadata object.

```{r}
#| label: add-rctd-full-to-metadata

# Add cell type weights to metadata
visium <- AddMetaData(visium, rctd_full@results$weights)

# Check this has been added as new columns to metadata
colnames(visium[[]])
```

We now have one column for each cell type, with the respective weight assigned by RCTD.
You can use these columns to visualise in spatial plots and UMAPs just as you would any other column.

### Identify most abundant type

We can also add a column metadata that determines which is the most abundant cell type in each spot.
We go back to our long format data frame of weights, and for each spot identify the cell type with the highest weight.
Then, we add this information back to the Seurat metadata.

```{r}
#| label: add-max-type-to-metadata

max_cell_type <- rctd_weights |>
  # for each barcode
  group_by(spot_barcode) |>
  # get the cell type with the highest weight
  slice_max(weight) |>
  ungroup()

# add to metadata - make sure to match the order of the barcodes
visium$rctd_max_type <- max_cell_type$subclass[match(
  colnames(visium),
  max_cell_type$spot_barcode
)]
visium$rctd_max_weight <- max_cell_type$weight[match(
  colnames(visium),
  max_cell_type$spot_barcode
)]

# Visualise the most abundant cell type in each spot
SpatialDimPlot(visium, group.by = "rctd_max_type") +
  ggtitle("Most abundant cell type in each spot")
```

You could also add a threshold to only assign a cell type if the weight is above a certain value, for example > 0.5.

```{r}
#| label: add-threshold-to-max-type

# add metadata column indicating whether the max weight is greater than 0.5
visium$rctd_max_type_gt_05 <- ifelse(
  visium$rctd_max_weight > 0.5,
  visium$rctd_max_type,
  NA
)

# Visualise the most abundant cell type with weight > 0.5
SpatialDimPlot(visium, group.by = "rctd_max_type_gt_05") +
  ggtitle("Most abundant cell type in each spot (weight > 0.5)")
```

As we can see, very few spots have cells with > 50% abundance, suggesting that most spots contain a mixture of cell types, which might be expected for Visium v1 data.
It may also be the case that, as we're running the analysis in full mode, the model is assigning small weights to many cell types, and therefore the maximum weight for each spot is relatively low.

You can explore these thresholds to explore the deconvolution results in different ways, and to identify the most likely cell type in each spot based on the weights assigned by RCTD.

### Weight Heatmaps

Finally, we can visualise the proportions of each cell type in each spot as a heatmap.
We could use information about our prior clustering to order the spots on the heatmap, for example.
Here, we order the spots based on the Leiden clusters we identified in the previous chapter, which are stored in the `Leiden_08` column of the metadata.

```{r}
#| label: heatmap-weights

# prepare heatmap matrix from weights matrix
# order the rows (spots) based on Leiden clusters
heatmap_data <- rctd_full@results$weights[order(visium$Leiden_08), ]

# Transpose the matrix for easier heatmap interpreation (cell types as rows, spots as columns)
heatmap_data <- t(heatmap_data)

# Visualise as heatmap adding leiden cluster annotation
pheatmap(
  heatmap_data,
  cluster_rows = FALSE,
  cluster_cols = FALSE,
  show_colnames = FALSE,
  annotation_col = data.frame(Leiden_08 = sort(visium$Leiden_08)),
  main = "Cell Type Proportions"
)
```

We can see some correlation between the Leiden clusters and the cell type proportions, which is expected as both are based on gene expression profiles.
This is not perfect however, we can also see many scattered spots with high proportions of certain cell types that do not correlate with these clusters.
This is not unexpected, as we know that each spot likely contains multiple cell types, and the clustering is based on the overall gene expression profile of the spot, which can be influenced by the presence of multiple cell types.

Alternatively, we could also cluster the columns of the heatmap to identify groups of spots with similar cell type compositions.

```{r}
#| label: heatmap-cluster-columns

# Heatmap with clustered spots
pheatmap(
  heatmap_data,
  cluster_rows = FALSE,
  cluster_cols = TRUE,
  show_colnames = FALSE,
  annotation_col = data.frame(Leiden_08 = sort(visium$Leiden_08)),
  main = "Cell Type Proportions"
)
```

We can see that the spots are clustered into groups with similar cell type compositions, which can be useful to identify spatial domains with similar cellular compositions.

For example, there is a very small cluster of spots with high proportions of "L6b":

```{r}
#| label: spatial-L6b
#| code-fold: true

# Spatial plot for L6b weight
spatial_l6b <- SpatialFeaturePlot(visium, features = "L6b") +
  ggtitle("L6b cell type")

# UMAP with these cells highlighted
umap_l6b <- DimPlot(
  visium,
  reduction = "umap",
  cells.highlight = WhichCells(visium, expression = rctd_max_type == "L6b")
) +
  theme(legend.position = "none") +
  coord_equal() +
  theme_void()

spatial_l6b + umap_l6b
```

These form a very narrow region in our tissue, in fact often spanning a single line of spots.
This is consistent with the fact that layer 6b neurons are a very specific cell type located in a narrow layer of the cortex, and therefore we would expect to find them in a small number of spots in our data.

This also illustrates the importance of resolution when analysing spatial data.
For example, in our previous [spatial clustering chapter](08-TissueArchitecture.qmd), we identified spatial clusters using BANKSY.
We used `k_geom = 18` for our analysis, which means that spatial expression features were averaged across 18 neighbours for each spot.
As we can see, for a narrow region like this, this averaging across neighbours can lead to the signal being diluted and not being identified as a distinct spatial domain, which is what we observed in our previous analysis.

This again highlights the importance of **biological knowledge of your tissue** and **considering the resolution of your data**.

## Exporting Seurat Object

Optionally, we can save the Seurat object with the deconvolution results added to the metadata for future use.

```{r}
#| label: save-seurat-object
#| eval: false

# Save Seurat object
saveRDS(visium, "results/mouse_brain_visium_deconvolution.rds")
```

```{r}
#| label: save-participant-data
#| include: false

saveRDS(visium, "precomputed/mouse_brain_visium_deconvolution.rds")
```

## Summary

::: {.callout-tip}
#### Key Points

- Deconvolution is a computational technique used to estimate the cellular composition of complex tissues from bulk gene expression data.
- The `RCTD` package provides functions for deconvolution of spatial transcriptomics data.
- Preparing both single-cell reference data and spatial transcriptomics data is essential for successful deconvolution analysis.
- Parallel processing can significantly speed up the computation time for deconvolution.
:::
