# Spatially Variable Features

```{r}
#| label: setup
#| include: false

knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)

# set the working directory to the course_files folder
# make sure to run course_files/download_course_files.sh
knitr::opts_knit$set(root.dir = here::here("course_files"))
```

::: {.callout-tip}
#### Learning Objectives

- Identify spatially variable features in spatial transcriptomics data
- Visualise spatially variable features on spatial maps
- Understand the biological significance of spatially variable features
:::

## Setup

We continue working with the sagittal mouse brain dataset from previous chapters.

::: {.callout-note collapse="true"}
#### Click to expand

Start by loading the required libraries and the Seurat object.

```{r}
#| label: load-visium-data

# Load libraries
library(Seurat) # single-cell and spatial analysis toolkit
library(sparseMatrixStats)
library(paletteer) # colour palettes
library(ggplot2) # plotting
library(dplyr) # data manipulation
library(patchwork) # combining plots

# Load the Seurat object from the previous chapter
visium <- readRDS("precomputed/mouse_brain_clustering.rds")
```

As a reminder, this object was created by:

- Importing the raw data using `Load10X_Spatial()`, followed by QC and filtering to remove low-quality spots.
- Data normalisation using `SCTransform()`, which models technical variation and applies variance stabilisation.
  Stored in the "SCT" assay.
- Applying dimensionality reduction using PCA, UMAP and t-SNE.
- Clustering the data using graph-based clustering.
  The clusters we use in this chapter were generated using the Leiden algorithm at a resolution of 0.8 and were set the the default cell identities (`Idents(visium)`).
:::

## Overview

In the [dimensionality reduction chapter](03-DimensionalityReduction.qmd) we covered how to perform feature selection of genes by looking for highly variable genes (HVGs).
The approach we used there was non-spatial, i.e. it assumed that spots are independent from each other, ignoring their spatial proximity in the tissue.

An alternative approach to feature selection is to use spatially-aware methods for finding variable genes.
These methods identify **spatially variable genes (SVGs)**, showing significant variation in expression across different spatial locations within a tissue sample.

Identifying these features serves a couple of different goals:

- It can complement [cluster marker gene analysis](04b-ClusterMarkerGenes.qmd), by providing additional genes that contribute to the spatial organisation of gene expression and the underlying biological processes.
- They can be used as input to performing a PCA, indirectly providing a spatial component to downstream analyses that rely on the PCA output (e.g. UMAP, t-SNE, graph-based clustering).

Methods for identifying spatially variable expression are often based on algorithms commonly used with geographic data.
Two common methods used are:

- **Moran's I**: Measures the correlation of expression between a point and the average around it. This average is calculated in a weighted manner, such that points closer to the focal point contribute more than points further away, giving this metric its spatial component.
  - Moran's I varies between -1 and 1.
    - Negative values represent dispersed patterns of expression. E.g. several distinct regions expressing the gene.
    - Positive values indicate clustering more concentrated in particular regions. E.g. increased expression at one end of the tissue.
    - A value of zero represents randomness in spatial distribution.

- **Markvariogram**: Measures how the differences in expression between pairs of points changes as a function of their distance.
  - A gene is spatially variable if the variance of expression between nearby points is significantly different from the variance between distant points.
  - This method can capture more complex spatial patterns than Moran's I, such as gradients or periodic patterns.

## Identifying SVGs

Seurat provides the `FindSpatiallyVariableFeatures()` function to identify SVGs.
This function performs statistical tests based on either Moran's I or the markvariogram method (default), to identify genes that exhibit significant spatial variability.

Here, we use it to identify the top 100 spatially variable features using Moran's I:

```{r}
#| label: find-svg-moransi

# Identify spatially variable features using the "moransi" method
visium <- FindSpatiallyVariableFeatures(
  visium,
  assay = "SCT",
  selection.method = "moransi",
  nfeatures = 2000
)

# View the top spatially variable features
head(SpatiallyVariableFeatures(visium), 10)
```

If you would like to compare the results of both methods, you will have to run the markvariogram method on a copy of the Seurat object, as running `FindSpatiallyVariableFeatures` again will overwrite the previous results.

As usual, it is useful to visualise these genes on spatial plots.
We will also add the spatial plot with our clustering results, to see how the spatially variable features relate to them.

```{r}
#| label: visualise-svg

# Visualise the top nine spatially variable features for a 3x3 grid
SpatialFeaturePlot(
  visium,
  features = SpatiallyVariableFeatures(visium)[1:9],
  ncol = 3
)

# Visualise the Leiden clusters (default Idents)
SpatialDimPlot(
  visium,
  label = TRUE,
  label.size = 3
) +
  theme(legend.position = "none")
```

We can see that some, but not all, of the spatially variable genes show spatial patterns matching our previously computed clusters.
While those clusters are not spatially aware (graph-based Leiden clusters), they still capture some of the spatial variation in the data, potentially because some cell types are spatially organised.

We will cover spatially aware clustering in a [later section](08-TissueArchitecture.qmd) and compare the results of those clusters and their marker genes to the spatially variable features we have identified here.

:::{.callout-tip}
#### Biological interpretation of spatially variable genes

Spatially variable genes help uncover biological processes and cellular identities in the tissue.
For example, genes that are spatially variable may be involved in cell-cell communication, tissue organisation, or responses to environmental cues.
Similarly to our earlier analysis of cluster marker genes, the biological significance of these spatially variable genes could be explored with pathway enrichment analysis or gene ontology analysis.
:::

## Summary

::: {.callout-tip}
#### Key Points

- Spatially variable features are genes or transcripts that show significant variation in expression across different spatial locations within a tissue sample.
- The `FindSpatiallyVariableFeatures` function in Seurat can be used to identify spatially variable features using various statistical methods.
- Visualizing spatially variable features on spatial maps can help to identify spatial patterns and relationships with other features such as clusters.
- Spatially variable features can provide insights into the biological processes and cellular interactions occurring within a tissue.
:::
