# Spatial Clustering

```{r}
#| label: setup
#| include: false

knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)

# set the working directory to the course_files folder
# make sure to run course_files/download_course_files.sh
knitr::opts_knit$set(root.dir = here::here("course_files"))
```

::: {.callout-tip}
### Learning Objectives

- Use BANKSY to augment gene expression data with spatial features and perform spatially-aware clustering.
- Interpret BANKSY clusters and explain how they differ from non-spatial clustering results.
- Identify relationships between BANKSY clusters, spatially variable genes, and cell type identities.
- Adjust BANKSY parameters (`lambda` and `k_geom`) to control the balance between spatial coherence and transcriptional similarity.
:::

## Setup

We continue working with the sagittal mouse brain dataset from previous chapters.

::: {.callout-note collapse="true"}
#### Click to expand

Start by loading the required libraries and the Seurat object.

```{r}
#| label: load-visium-data

# Load libraries
library(Seurat) # single-cell and spatial analysis toolkit
library(sparseMatrixStats)
library(paletteer) # colour palettes
library(ggplot2) # plotting
library(dplyr) # data manipulation
library(patchwork) # combining plots

# Load the Seurat object from the previous chapter
visium <- readRDS("precomputed/mouse_brain_svgs.rds")
```

As a reminder, this object was created by:

- Importing the raw data using `Load10X_Spatial()`, followed by QC and filtering to remove low-quality spots.
- Data normalisation using `SCTransform()`, which models technical variation and applies variance stabilisation.
  Stored in the "SCT" assay.
- Applying dimensionality reduction using PCA, UMAP and t-SNE.
- Clustering the data using graph-based clustering.
  The clusters we use in this chapter were generated using the Leiden algorithm at a resolution of 0.8 and were set the the default cell identities (`Idents(visium)`).
- Identifying spatially variable features using Moran's I and markvariogram methods.
:::

## Overview

In our [previous clustering chapter](04-Clustering.qmd), we used graph-based clustering to group spots based solely on their transcriptional information, without considering spatial organisation.

**Tissue architecture** refers to the spatial organisation and arrangement of cell types within tissue.
To capture this, we need methods that incorporate spatial information into the clustering process.

BANKSY is a spatially aware clustering method that combines gene expression with spatial context.
Unlike standard graph-based clustering, BANKSY first augments the gene expression matrix by adding spatial features, then applies clustering to this enriched representation.

For each gene, BANKSY calculates two spatial features:

- **Mean neighbourhood expression**: The average expression of a gene across neighbouring spots, weighted so nearby spots contribute more strongly than distant ones.
- **Azimuthal Gabor filter (AGF)**: Inspired by image-processing methods, this captures how gene expression changes across a spot's neighbourhood by measuring directional differences in expression.
  Intuitively, it identifies directional changes in expression and boundaries (e.g. the "gradient" of expression) highlighting spots that differ from their surrounding tissue environment.

These spatial features are combined with the original gene expression values before clustering.
The result is that BANKSY identifies tissue domains defined by both gene expression and spatial expression patterns.

## Running BANKSY

BANKSY is built on Bioconductor's `SpatialExperiment` class, but the `SeuratWrappers` package provides a convenient interface to run BANKSY directly on Seurat objects.
Let us load these packages:

```{r}
#| label: load-banksy

# Load necessary libraries
library(SeuratWrappers)
library(Banksy)
```

The `RunBanksy()` function accepts two key parameters that control clustering behaviour:

- **`lambda`**: Controls the balance between spatial and transcriptional information.
  - `lambda = 0`: uses only transcriptional information (equivalent to standard graph-based clustering)
  - `lambda = 1`: uses only spatial information from the neighbourhood
  - Values in between give a weighted balance, allowing you to capture both spatial domains and cell identities

- **`k_geom`**: Controls the number of spatial neighbours used to calculate mean neighbourhood expression and the azimuthal Gabor filter.
  A larger value includes more distant neighbours.

By default, `RunBanksy()` does not calculate AGF features (as it is computationally intensive), but you can enable this with `use_agf = TRUE`.
We will use all genes for the analysis to keep it independent of previous variable feature selection.

```{r}
#| label: precompute-banksy
#| echo: false

# To save time, create/read precomputed object
if (!file.exists("precomputed/mouse_brain_banksy.rds")) {
  visium <- RunBanksy(
    visium,
    assay = "SCT",
    lambda = 0.2,
    k_geom = 15,
    use_agf = TRUE,
    features = "all"
  )
  saveRDS(visium, "precomputed/mouse_brain_banksy.rds")
} else {
  visium <- readRDS("precomputed/mouse_brain_banksy.rds")
}

# Confirm that the BANKSY assay has been added to the Seurat object
Assays(visium)

# Check the layers in the BANKSY assay
Layers(visium[["BANKSY"]])
```

```{r}
#| label: run-banksy
#| eval: false

# Run BANKSY on the Seurat object
visium <- RunBanksy(
  visium,
  assay = "SCT",
  lambda = 0.2,
  k_geom = 15,
  use_agf = TRUE,
  features = "all"
)

# Confirm that the BANKSY assay has been added to the Seurat object
Assays(visium)

# Check the layers in the BANKSY assay
Layers(visium[["BANKSY"]])
```

`RunBanksy()` creates a new assay (by default named "BANKSY") containing the augmented gene expression matrix with a `scale.data` layer ready for downstream analysis.
Note that this matrix has three times the number of features as the original assay because it includes the original gene expression, mean neighbourhood expression, and AGF features for each gene.

```{r}
#| label: banksy-nrow

# Compare the dimensions of the SCT and BANKSY assays
nrow(visium[["SCT"]])
nrow(visium[["BANKSY"]])
```

### Clustering

With the augmented BANKSY gene expression, we proceed to cluster the data using the standard workflow:

1. Run PCA on the BANKSY assay to reduce dimensionality.
2. Find nearest neighbours based on the PCA reduction.
3. Cluster the data using graph-based clustering (Leiden or Louvain).

We use 50 principal components and a resolution of 0.8 for the Leiden algorithm, matching our previous analysis.
We store the new cluster identities in a column called "banksy_cluster" to distinguish them from the earlier non-spatial clusters.

```{r}
#| label: banksy-clustering

visium <- visium |>
  RunPCA(
    assay = "BANKSY",
    reduction.name = "pca_banksy",
    reduction.key = "pcabanksy_",
    features = rownames(visium[["BANKSY"]]),
    npcs = 50
  ) |>
  FindNeighbors(reduction = "pca_banksy", dims = 1:50) |>
  FindClusters(cluster.name = "banksy_cluster", resolution = 0.8)
```

### Visualisation

We next visualise the BANKSY clusters on both UMAP and spatial plots to see how they relate to tissue organisation.
For comparison, we also display the earlier non-spatial clusters from the SCT assay.

```{r}
#| label: banksy-visualisation
#| code-fold: true
#| code-summary: "Show code for BANKSY cluster visualisation"

# Run UMAP on the BANKSY PCA reduction
visium <- RunUMAP(
  visium,
  reduction = "pca_banksy",
  dims = 1:50,
  reduction.name = "umap_banksy"
)

# UMAP plot
banksy_umap <- DimPlot(
  visium,
  reduction = "umap_banksy",
  group.by = "banksy_cluster",
  label = TRUE
) +
  ggtitle("BANKSY Clusters: lambda = 0.2, k_geom = 15") +
  coord_equal() +
  theme_void()

# Spatial plot
banksky_spatial <- SpatialDimPlot(
  visium,
  group.by = "banksy_cluster",
  label = TRUE,
  label.size = 4,
  image.alpha = 0.5
) +
  theme(legend.position = "none")

# SCT assay UMAP and spatial plots for comparison
sct_umap <- DimPlot(
  visium,
  reduction = "umap",
  group.by = "Leiden_08",
  label = TRUE
) +
  ggtitle("SCT Clusters") +
  coord_equal() +
  theme_void()

sct_spatial <- SpatialDimPlot(
  visium,
  group.by = "Leiden_08",
  label = TRUE,
  label.size = 4,
  image.alpha = 0.5
) +
  theme(legend.position = "none")

# Plot side by side
(banksy_umap + banksky_spatial) /
  (sct_umap + sct_spatial)
```

The BANKSY UMAP differs from the SCT UMAP because the BANKSY reduction incorporates spatial information.
Notice how the clusters show much less mixing in both the UMAP and spatial plots, indicating tighter spatial coherence.
An illustrative example is SCT Cluster 5, which is split across several separate regions of the tissue.
BANKSY fragments this scattered cluster into more spatially unified domains.

```{r}
#| label: highlight-sct-cluster
#| code-fold: true
#| fig.width: 5
#| fig.height: 5

# Highlight a particular cluster of interest
sct_c5 <- SpatialPlot(
  visium,
  cells.highlight = WhichCells(visium, expression = Leiden_08 == "5"),
  image.alpha = 0.5
) +
  ggtitle("SCT cluster 5") +
  theme(legend.position = "none")
```

This visualisation shows how BANKSY emphasises tissue architecture.
Even with a relatively modest lambda value of 0.2, the spatial structure is clearly visible.

It is important to understand that BANKSY clusters are not "better" than non-spatial clusters - they answer different biological questions.
Non-spatial clustering identifies **cell identities**: groups of cells with similar transcriptional profiles, regardless of location.
The same cell type may appear in different tissue regions.
BANKSY identifies **tissue domains**: spatial regions defined by both transcriptional profiles and spatial context.
A tissue domain may contain multiple cell types but represents a coherent spatial unit.

::: {.callout-important collapse="true"}
#### Important note on the choice of lambda value in Visium V1 data

We copy a warning directly from the [BANKSY documentation](https://prabhakarlab.github.io/Banksy/), as it is relevant to our dataset:

> **An important note about choosing the `lambda` parameter for the older [Visium v1 / v2 55um > datasets](https://doi.org/10.1038/s41593-020-00787-0) or the original [ST 100um technology](https://doi.org/10.1038/s41596-018-0045-2):**
>
> For most modern high resolution technologies like Xenium, Visium HD, StereoSeq, MERFISH, STARmap PLUS, SeqFISH+, SlideSeq v2, and CosMx (and others), we recommend the usual defults for `lambda`: For cell typing, use `lambda = 0.2` (as shown below, or in [this > vignette](https://prabhakarlab.github.io/Banksy/articles/parameter-selection.html)) and for [domain > segmentation](https://prabhakarlab.github.io/Banksy/articles/domain-segment.html), use `lambda = 0.8`.
> These technologies are either imaging based, having true single-cell resolution (e.g., MERFISH), or are sequencing based, having barcoded spots on the scale of single-cells (e.g., [Visium > HD](https://www.10xgenomics.com/products/visium-hd-spatial-gene-expression)).
> We find that the usual defaults work well at this measurement resolution.
>
> However, for the older **Visium v1/v2** or **ST** technologies, with their much lower resolution spots (55um and 100um diameter, respectively), we find that `lambda = 0.2` seems to work best for domain segmentation.
> This could be because each spot already measures the average transcriptome of several cells in a neighbourhood.
> It seems that `lambda = 0.2` shares enough information between these neighbourhoods to lead to good domain segmentation performance.
> For example, in the [human DLPFC > vignette](https://prabhakarlab.github.io/Banksy/articles/multi-sample.html), we use `lambda = 0.2` on a Visium v1/v2 dataset.
> Also note that in these lower resolution technologies, each spot can have multiple cells of different types, and as such *cell-typing* is not defined for them.
:::

## Comparison with SVGs

We can identify differentially expressed genes between BANKSY clusters using `FindAllMarkers()`.
Because the BANKSY clusters are spatially informed, their marker genes indirectly reflect tissue architecture.

However, which assay we use matters:

- Use the "SCT" assay to identify genes differentially expressed between BANKSY clusters in the original gene expression space.
- If we use the "BANKSY" assay, we would identify genes that are differentially expressed in the augmented space including spatial features.
  This may not directly reflect gene expression differences.

We therefore use the "SCT" assay:

```{r}
#| label: banksy-markers

# Find markers for BANKSY clusters
banksy_markers <- FindAllMarkers(
  visium,
  assay = "SCT",
  group.by = "banksy_cluster",
  only.pos = TRUE,
  min.pct = 0.25,
  logfc.threshold = 0.25
)

# Take all genes with adjusted p-value < 0.05 and 2x fold change
banksy_marker_genes <- banksy_markers |>
  filter(p_val_adj < 0.05 & avg_log2FC > 1) |>
  distinct(gene) |>
  pull(gene)
```

This returns `r length(banksy_marker_genes)` marker genes for the BANKSY clusters.

Now we compare these BANKSY marker genes with the 2000 spatially variable genes (SVGs) identified in the previous chapter:

```{r}
#| label: marker-overlap

# Get the 2000 spatially variable genes from previous analysis
svg_genes <- SpatiallyVariableFeatures(visium[["SCT"]])

# Quick Venn diagram
gplots::venn(list(
  BANKSY = banksy_marker_genes,
  SVGs = svg_genes,
  Leiden = hvg_genes
))
```

Approximately 70% of the previous SVGs are also identified as BANKSY marker genes, but many BANKSY marker genes are not among the SVGs.

Again, we emphasise that this doesn't reflect one method being better than the other.
SVGs are selected based on spatial variability across the tissue - high expression in some locations, low in others.
BANKSY markers are selected based on differential expression between spatially informed clusters.

The overlap suggests that BANKSY clusters capture genuine spatial patterns in gene expression.
The additional BANKSY markers not identified as SVGs may reflect cell identity markers within spatially defined domains, or fine-scale spatial patterns not captured by SVG methods.

Conversely, non-spatial clustering may identify cell type clusters that are not spatially organised or sparsely distributed across regions (for example, scattered immune cells in tumour tissue), which BANKSY might miss because its focus on spatial coherence would merge them with surrounding tissue.

## Exercises

::: {.callout-exercise}
#### Exploring lambda

In the code above, we ran BANKSY with a lambda value of 0.2, which weights transcriptional information more heavily than spatial information.
Try running BANKSY with a higher lambda value of 0.8, which weights spatial information more heavily, and compare the results to the original analysis.

To avoid overwriting the previous results:

- In `RunBanksy()` use the `assay_name` option to name the new BANKSY assay differently (e.g. "BANKSY_lambda_0.8").
- In `FindClusters()` use the `cluster.name` option with a different name (e.g. "banksy_cluster_lambda_0.8").

::::: {.callout-hint}
As a reminder, here is the complete workflow we ran earlier:

```{r}
#| label: banksy-hint
#| eval: false

# Run BANKSY
visium <- visium |>
  RunBanksy(
    assay = "SCT",
    lambda = 0.2,
    k_geom = 15,
    use_agf = TRUE,
    features = "all"
  )

# PCA, Clustering and UMAP
visium <- visium |>
  RunPCA(
    assay = "BANKSY",
    reduction.name = "pca_banksy",
    reduction.key = "pcabanksy_",
    features = rownames(visium[["BANKSY"]]),
    npcs = 50
  ) |>
  FindNeighbors(reduction = "pca_banksy", dims = 1:50) |>
  FindClusters(cluster.name = "banksy_cluster", resolution = 0.8) |>
  RunUMAP(
    reduction = "pca_banksy",
    dims = 1:50,
    reduction.name = "umap_banksy"
  )
```

You can adapt this workflow by changing the lambda value and assay and cluster names.
:::::

::::: {.callout-answer}
We adapt the workflow with the new lambda value and updated assay and cluster names.
Note that the names are intentionally verbose for clarity, but you may choose to be more concise in your own analysis.

```{r}
#| label: banksy-lambda-0.8

# Run BANKSY
visium <- visium |>
  RunBanksy(
    assay = "SCT",
    lambda = 0.8,
    k_geom = 15,
    use_agf = TRUE,
    features = "all",
    assay_name = "banksy_lambda_08"
  )

# PCA, Clustering and UMAP
visium <- visium |>
  RunPCA(
    assay = "banksy_lambda_08",
    reduction.name = "pca_banksy_lambda_08",
    reduction.key = "pcabanksylambda08_",
    features = rownames(visium[["banksy_lambda_08"]]),
    npcs = 50
  ) |>
  FindNeighbors(reduction = "pca_banksy_lambda_08", dims = 1:50) |>
  FindClusters(cluster.name = "banksy_cluster_lambda_08", resolution = 0.8) |>
  RunUMAP(
    reduction = "pca_banksy_lambda_08",
    dims = 1:50,
    reduction.name = "umap_banksy_lambda_08"
  )

# Confirm the new assay and dimensionality reductions are present
Assays(visium)
Reductions(visium)
```

We can now plot these results alongside the lambda = 0.2 analysis:

```{r}
#| label: compare-lambda

# UMAP plot
banksy_umap_08 <- DimPlot(
  visium,
  reduction = "umap_banksy_lambda_08",
  group.by = "banksy_cluster_lambda_08",
  label = TRUE
) +
  ggtitle("lambda = 0.8, k_geom = 15") +
  coord_equal() +
  theme_void()

# Spatial plot
banksky_spatial_08 <- SpatialDimPlot(
  visium,
  group.by = "banksy_cluster_lambda_08",
  label = TRUE,
  label.size = 4,
  image.alpha = 0.5
) +
  theme(legend.position = "none")

# Plot alongside the previous lambda 0.2 results
(banksy_umap_08 + banksky_spatial_08) /
  (banksy_umap + banksky_spatial)
```

Two differences are apparent when comparing the analyses:

- The lambda = 0.8 analysis identifies fewer clusters than lambda = 0.2, because the clustering now prioritises spatial coherence over transcriptional similarity.
  Neighbouring tissue regions are more likely to be grouped together.
- If you rotate the lambda = 0.8 UMAP 180 degrees, it nearly recreates the original spatial tissue architecture.
  This demonstrates that the higher lambda value drives clustering primarily by spatial information, producing clusters that closely reflect underlying tissue organisation.

**The choice of lambda depends on your biological question.** For identifying tissue domains defined by spatial context, use a higher lambda value.
For identifying cell types that may be scattered across different regions, use a lower lambda value.
:::::
:::

::: {.callout-exercise}
#### Exploring neighbourhood size

The parameter `k_geom` controls the number of nearest neighbours used to calculate the spatially-augmented features by BANKSY.
A larger value will likely result in a more connected graph and amplifies spatial information.

Try running BANKSY with a higher `k_geom` value of 50, keeping lambda at 0.2 as in the original analysis.
Compare the results to the original analysis (lambda = 0.2, k_geom = 15).

::::: {.callout-answer}
```{r}
#| label: banksy-k-50

# Run BANKSY
visium <- visium |>
  RunBanksy(
    assay = "SCT",
    lambda = 0.2,
    k_geom = 50,
    use_agf = TRUE,
    features = "all",
    assay_name = "banksy_k_50"
  )

# PCA, Clustering and UMAP
visium <- visium |>
  RunPCA(
    assay = "banksy_k_50",
    reduction.name = "pca_banksy_k_50",
    reduction.key = "pcabanksyk50_",
    features = rownames(visium[["banksy_k_50"]]),
    npcs = 50
  ) |>
  FindNeighbors(reduction = "pca_banksy_k_50", dims = 1:50) |>
  FindClusters(cluster.name = "banksy_cluster_k_50", resolution = 0.8) |>
  RunUMAP(
    reduction = "pca_banksy_k_50",
    dims = 1:50,
    reduction.name = "umap_banksy_k_50"
  )

# Confirm the new assay and dimensionality reductions are present
Assays(visium)
Reductions(visium)
```

We create UMAP and spatial plots for this analysis and compare them to the original:

```{r}
#| label: compare-k

# UMAP plot
banksy_umap_k50 <- DimPlot(
  visium,
  reduction = "umap_banksy_k_50",
  group.by = "banksy_cluster_k_50",
  label = TRUE
) +
  ggtitle("Lambda = 0.2, k_geom = 50") +
  coord_equal() +
  theme_void()

# Spatial plot
banksky_spatial_k50 <- SpatialDimPlot(
  visium,
  group.by = "banksy_cluster_k_50",
  label = TRUE,
  label.size = 4,
  image.alpha = 0.5
) +
  theme(legend.position = "none")

# Plot alongside the previous results
(banksy_umap_k50 + banksky_spatial_k50) /
  (banksy_umap + banksky_spatial)
```

Increasing `k_geom` to 50 seems to have two effects:

- Fewer clusters are identified because the larger neighbourhood size creates a more connected graph, which merges smaller spatial regions into larger domains.
- The UMAP more closely reflects the original tissue architecture, because the larger spatial neighbourhood emphasises spatial coherence into the clustering.

The choice of `k_geom` reflects your prior assumption about tissue domain size.
Lower values are appropriate for data with finely structured tissue.
Higher values smooth over fine structure and identify larger domains.

For Visium v1/v2 data (like our dataset), each spot already measures the average expression from multiple cells occupying that location.
This implicitly incorporates spatial information.
You may therefore want to use lower `k_geom` and `lambda` values than for high-resolution technologies such as Visium HD and Xenium.
:::::
:::

## Summary

::: {.callout-tip}
### Key Points

- Use `RunBanksy()` to augment expression data with spatial features: mean neighbourhood expression (average expression in neighbouring spots) and azimuthal Gabor filter (directional changes in expression).
  - `RunBanksy()` accepts parameters: `lambda` (balance between spatial and transcriptional information), `k_geom` (number of spatial neighbours), and `use_agf` (whether to calculate AGF features).
  - The resulting augmented matrix has three times as many features as the original because it includes original expression plus two spatial features per gene.

- BANKSY clusters differ from non-spatial clusters because they prioritise spatial coherence in addition to transcriptional similarity.
  - Non-spatial clustering identifies **cell identities** — cells with similar expression profiles, regardless of location.
  - BANKSY identifies **tissue domains** — spatially coherent regions defined by both expression and spatial context, which may contain multiple cell types.
  - Visualising both methods side-by-side reveals how BANKSY fragments scattered clusters into spatially unified regions.

- BANKSY marker genes overlap substantially with spatially variable genes (SVGs) but include additional markers reflecting cell identity within spatial domains.
  - ~70% of SVGs are recovered as BANKSY marker genes, indicating BANKSY captures spatial variability.
  - BANKSY markers not identified as SVGs may reflect uniform expression within a domain or cell type identity within a spatial region.
  - The complementary nature of these gene sets means non-spatial clustering and spatial analysis together provide different biological insights.

- Adjust lambda and k_geom to control the weighting between spatial coherence and transcriptional similarity.
  - Higher `lambda` (e.g. 0.8) emphasises spatial coherence, identifying larger tissue domains.
  - Lower `lambda` (e.g. 0.2) emphasises transcriptional similarity, allowing cell types to be identified across multiple spatial regions.
  - Higher `k_geom` increases neighbourhood size and spatial smoothing, resulting in fewer and larger clusters.
  - Lower `k_geom` creates more fragmented neighbourhoods, identifying finer tissue structure.
  - For Visium v1/v2 data, lower values work well because each spot already averages multiple cells' expression.
:::
