# Cell-Cell Communication

```{r}
#| label: setup
#| include: false

knitr::opts_chunk$set(
  echo = TRUE,
  message = FALSE,
  warning = FALSE,
  cache.lazy = FALSE
)

# set the working directory to the course_files folder
# make sure to run course_files/download_course_files.sh
knitr::opts_knit$set(root.dir = here::here("course_files"))
```

::: {.callout-tip}
#### Learning Objectives

- Understand the concept of cell-cell interactions in spatial transcriptomics
- Learn how to analyze cell-cell interactions using the `CellChat` package
- Interpret the results of cell-cell interaction analysis
- Explore different methods for visualizing cell-cell interactions
:::

## Setup

We continue working with the sagittal mouse brain dataset from previous chapters.

::: {.callout-note collapse="true"}
#### Click to expand

Start by loading the required libraries and the Seurat object.

```{r}
#| label: load-visium-data

# Load libraries
library(Seurat) # single-cell and spatial analysis toolkit
library(sparseMatrixStats)
library(paletteer) # colour palettes
library(ggplot2) # plotting
library(dplyr) # data manipulation
library(patchwork) # combining plots

# Load the Seurat object from the previous chapter
visium <- readRDS("precomputed/mouse_brain_visium_deconvolution.rds")
```

As a reminder, this object was created by:

- Importing the raw data using `Load10X_Spatial()`, followed by QC and filtering to remove low-quality spots.
- Data normalisation using `SCTransform()`, which models technical variation and applies variance stabilisation.
  Stored in the "SCT" assay.
- Applying dimensionality reduction using PCA, UMAP and t-SNE.
- Clustering the data using graph-based clustering.
  The clusters we use in this chapter were generated using the Leiden algorithm at a resolution of 0.8 and were set the the default cell identities (`Idents(visium)`).
- Identifying spatially variable features using Moran's I and markvariogram methods.
- Running the BANKSY algorithm to identify spatial domains in the data, which were stored in the `banksy_clusters` column of the metadata.
- Using the RCTD package to perform deconvolution of the spatial transcriptomics data, using a single-cell RNA-seq reference dataset of the mouse brain.
:::

## Overview

In multicellular tissues, cells communicate with one another to coordinate processes such as development, immune responses and disease progression.
One of the most common communication mechanisms involves **signalling molecules (ligands)** produced by one cell and detected by **receptors** expressed on another cell.
Receptor-mediated pathways can then effect gene expression, cellular behaviour and cell fate decisions.

Single-cell and spatial transcriptomics allow looking at these interactions based on gene expression across thousands of cells simultaneously.
Cell-cell communication analysis (CCC) uses **prior knowledge of ligand-receptor pairs** to identify potential signalling relationships between cell populations.
The general idea is that cells expressing a ligand are considered potential "senders", while cells expressing the corresponding receptor are considered potential "receivers".
This can be used to build an interaction network for generating hypotheses of how information flows through a tissue.

It is important to emphasise that **CCC methods rely on curated databases of known ligand-receptor interactions**.
Rather than discovering entirely new signalling relationships, these methods use observed gene expression patterns to identify which known interactions are most likely to occur between cell populations.

Spatial transcriptomics provides additional information beyond single-cell transcriptomics by preserving the physical location of cells within a tissue, allowing inferred interactions to be evaluated in the context of spatial proximity.
This is useful as many signalling mechanisms occur locally and are therefore more plausible when sender and receiver cells are located near one another.
This means that we can explore how cell types communicate and influence each other within the tissue microenvironment.

Many packages have been developed for analysing cell-cell communication (CCC) from single-cell and spatial transcriptomics data.
As is the case with many spatial analysis workflows, this is an active area of development, and we encourage you to regularly look for benchmark studies comparing current tools.
For example, [Ku et al. 2026](https://doi.org/10.1186/s13059-026-04063-5) benchmarked a range of spatially-aware and non-spatial CCC inference methods across multiple spatial transcriptomics technologies.

One of their main conclusions is that there is no universally best method for CCC inference across all spatial technologies.
Performance depends on factors such as spatial resolution, cell type composition and the underlying assumptions of the method.
In general, methods originally developed for single-cell data perform well on high-resolution spatial datasets, while spatially-aware methods can give complementary information for lower-resolution technologies.

We discuss more caveats and limitations of these methods at the end of this chapter.
For now, we will demonstrate how to perform CCC using a popular package developed for single-cell CCC inference: `CellChat`.
Note that the benchmark study above uses CellChat v2 (which we are also using here), but [v3 is under development](https://github.com/jinworks/SpatialCellChat) and provides further functionality for spatial data.
This emphasises the speed of development in the field, and why regularly looking for benchmark studies is important.

::: {.callout-tip}
#### Acronyms used in this chapter

Throughout this chapter we use the term **cell-cell communication (CCC)** to describe signalling relationships inferred from ligand-receptor expression patterns.
The literature also frequently uses the term **cell-cell interaction (CCI)**, and the two terms are often used interchangeably in the context of transcriptomics-based analyses.
:::

## CellChat Analysis

*CellChat* is an R package for inference of cell-cell communication (CCC) networks from single-cell and spatial transcriptomics data.
It combines gene expression measurements with a curated database of ligand-receptor interactions to estimate the probability of communication between cell populations.
It also accounts for multi-subunit receptor complexes, cofactors and pathway-level organisation, which improves its performance compared to methods that only use ligand-receptor pairs.

These are the steps required for the analysis:

- Create the CellChat object
- Set the ligand-receptor interaction database
- Idenfity overexpressed genes and interactions
- Compute communication probabilities and infer signaling pathways
- Visualise and explore the results

We will go through each of these steps in detail.
First, let's load the library:

```{r}
#| label: load-cellchat

# Load CellChat library
library(CellChat)
```

### Create CellChat Object

The CellChat object can be created directly from a Seurat object using the `createCellChat()` function.

Besides the Seurat object, this function requires a data frame of **spatial conversion factors** with two columns:

- **`ratio`**: the conversion factor for converting spatial coordinates from pixels to microns.
  For example a ratio of 0.1 would mean that 1 pixel = 0.1 μm.
- **`tol`**: a tolerance factor to allow flexibility when calculating interaction ranges between cells.
  The authors recommend setting this to half the spot diameter in microns.

The `ratio` factor can be calculated based on the spatial resolution of the data, as the ratio of the spot diameter in microns to the spot diameter in pixels.
The spot diameter in pixels is available as image metadata (remember you can list images in the Seurat object with `Images()`).

```{r}
#| label: view-scale-factors

# View scale factors for the image in the Seurat object
ScaleFactors(visium[["slice1"]])
```

The `spot` entry is what we're interesting in, and gives the number of pixels corresponding to the spot diameter (55 μm for 10X Visium v1/v2).

We can therefore create our spatial factors data frame as follows:

```{r}
#| label: create-spatial-factors

# Create spatial factors data frame
spatial_factors <- data.frame(
  ratio = 55 / ScaleFactors(visium[["slice1"]])$spot,
  tol = 55 / 2
)

spatial_factors
```

We also need a separate data frame with the spatial coordinates of each spot, which can be extracted from the Seurat object using the `GetTissueCoordinates()` function.

```{r}
#| label: extract-spatial-coordinates

# Extract tissue coordinates of the spots
spatial_coords <- GetTissueCoordinates(visium, image = "slice1")[, c("x", "y")]

head(spatial_coords)
```

Before creating the CellChat object, we can check the distances of the spots to each other, which will be used in the CellChat analysis.
We know that for 10X Visium v1 the minimum distance should be roughly 100 μm between the centers of adjacent spots, so this is a good check that our spatial factors are correct.

```{r}
#| label: check-spot-distances

# compute distances of the spots to each other
spot_dists <- computeCellDistance(
  coordinates = spatial_coords,
  ratio = spatial_factors$ratio,
  tol = spatial_factors$tol
)

# minimum distance observed
# remove zero distances (distance of a spot to itself)
min(spot_dists[spot_dists != 0])
```

This value is smaller than the 100 μm, but still within a reasonable range.

With our scale factors and spatial coordinates prepared, we can now create the CellChat object using the `createCellChat()` function.

```{r}
#| label: create-cellchat

# Create CellChat object
cell_chat <- createCellChat(
  object = visium,
  assay = "SCT",
  group.by = "first_type",
  datatype = "spatial",
  coordinates = spatial_coords,
  spatial.factors = spatial_factors
)

# drop levels of cell type labels that are not present in the data
cell_chat@idents <- droplevels(cell_chat@idents)
```

A few of the critical options we used include:

- `assay = "SCT"`: CellChat requires log-normalised expression data.
  We use the log-normalised data from the sctransform method ("SCT" assay).
  You could also specify `assay = "Spatial"` if you wanted to use the log-normalised data from the `LogNormalize` method, which is stored in that assay.
- `datatype = "spatial"`: this tells CellChat that we are working with spatial transcriptomics data, which allows it to use the spatial coordinates and factors in the analysis.
- `group.by = "first_type"`: this specifies the metadata column to use for cell type labels.
  We used the first cell type label from our deconvolution results (`first_type` column in the metadata).
  However, you could choose to use a different label, for example the `banksy_cluster` or the `Leiden_08` metadata columns from our spatial and non-spatial clustering analysis, respectively.
- We also provided the spatial coordinates and factors we prepared in the previous steps.

### Ligand-receptor database

One critical aspect of CCC analysis is the choice of ligand-receptor interaction database.
CellChat provides three built-in databases: `CellChatDB.mouse`, `CellChatDB.human` and `CellChatDB.zebrafish`.
For custom databases, you can use `updateCellChatDB()`, providing your own ligand-receptor interaction pairs (see the function's help for details).

Here, we use the mouse database, which contains `r nrow(CellChatDB.mouse$interaction)` ligand-receptor interactions.
We can quickly visualise the categories of ligand-receptor interactions available in this database:

```{r}
#| label: visualise-cellchat-db

# Visualise categories of interactions in the mouse database
showDatabaseCategory(CellChatDB.mouse)
```

Each category represents a different type of signaling mechanism that can occur between cells.
In the following code chunk, we will:

- Exclude the "Non-protein Signalling" category, which contains interactions that are not mediated by protein ligands and receptors.
- We add the database to the CellChat object using the `@DB` slot.
- We use the `subsetData()` function to subset the expression data of signalling genes.
  This reduces the computational cost of the analysis by only keeping the genes that are relevant for the interactions in the database subset.

```{r}
#| label: update-cellchat-db

# Subset database to exclude "Non-protein Signalling"
mouse_db <- subsetDB(
  CellChatDB.mouse,
  search = c("Secreted Signaling", "ECM-Receptor", "Cell-Cell Contact"),
  non_protein = FALSE
)

# add the database to the CellChat object
cell_chat@DB <- mouse_db

# subset the expression data of signaling genes for saving computation cost
cell_chat <- subsetData(cell_chat)
```

We now have `r nrow(cell_chat@data.signaling)` genes for cell-cell communication analysis.

### Variable Genes and Interactions

Before inferring the CCC network, CellChat identifies variable genes in the dataset.
This is effectively done as a differential expression test between the labels we set above (`first_type` in our case), similar to what we did in [cluster marker genes](04b-ClusterMarkerGenes.qmd).
It also identifies variable ligand-receptor interactions where both pairs in an interaction are variable.
These steps help to focus the analysis on the most relevant genes and interactions.

```{r}
#| label: identify-overexpressed-genes

# Identify overexpressed genes and interactions
cell_chat <- identifyOverExpressedGenes(cell_chat)
cell_chat <- identifyOverExpressedInteractions(cell_chat)
```

This identifies `r length(cell_chat@var.features$features)` variable genes and `r nrow(cell_chat@LR$LRsig)` variable ligand-receptor interactions.

### CCC Network Inference

The final step of the analysis is to compute the communication probabilities for each ligand-receptor interaction and infer the signaling pathways involved.

We start by computing the communication probability between cell groups using the `computeCommunProb()` function.

```{r}
#| label: compute-communication-probabilities
#| cache: true

# Compute communication probabilities
cell_chat <- computeCommunProb(
  cell_chat,
  contact.range = 100,
  interaction.range = 200
)
```

We set:

- `contact.range = 100` to restrict contact-dependent interactions to 100 micron radius.
  This is appropriate for 10X Visium data where the expected distance between the centers of two spots is around 100 microns.
  For single-cell resolution technologies the recommendation is to set this to the average cell diameter.
- `interaction.range = 200` to restrict secreted signaling interactions to 200 micron radius.
  This might be more dependent on your expectations of the biological plausibility of medium-range interactions in your tissue than the spatial resolution of your data.

We can also compute the communication probabilities at the signaling pathway level using the `computeCommunProbPathway()` function, which aggregates the probabilities of individual ligand-receptor interactions into pathway-level probabilities based on the known organization of interactions into pathways in the database.

```{r}
#| label: compute-pathway-communication-probabilities

# Compute communication probabilities at the signaling pathway level
cell_chat <- computeCommunProbPathway(cell_chat)
```

Finally, we can filter the inferred interactions to only consider those that are supported by a minimum number of cells.
We also aggregate the results of the computed network, which can be used in later visualisation steps.

```{r}
#| label: filter-communication

# Filter interactions to only consider those supported by at least 10 spots
cell_chat <- filterCommunication(cell_chat, min.cells = 10)

# Aggregate the communication network at the cell type level
cell_chat <- aggregateNet(cell_chat)
```

The resulting CellChat object now contains information about the identified interactions, their probabilities, and the overall communication network between different cell types in the dataset.

## CCC Network Visualisation

CellChat provides a range of functions for visualising the inferred cell-cell communication (CCC) network.

We start by visualising the overall communication network using a circle plot, which shows the number of interactions and the interaction weights/strength between different cell types.

```{r}
#| label: visualise-circle

# calculate number of cells of each type for scaling the node sizes in the plot
group_size <- as.numeric(table(cell_chat@idents))

# Visualise the overall communication network as a circle plot
netVisual_circle(
  cell_chat@net$count,
  vertex.weight = group_size,
  weight.scale = TRUE,
  label.edge = FALSE,
  title.name = "Number of interactions",
  top = 0.1
)

# Visualise the weights of the interactions
netVisual_circle(
  cell_chat@net$weight,
  vertex.weight = group_size,
  weight.scale = TRUE,
  label.edge = FALSE,
  title.name = "Interaction weights/strength",
  top = 0.1
)
```

The `top` option is used to show only the top 10% of interactions in the plot, but you can change this to show more or fewer interactions depending on how cluttered the plot looks.

We can also visualise the interaction strength as a heatmap:

```{r}
#| label: visualise-heatmap

# Visualise the interaction strength between cell types as a heatmap
netVisual_heatmap(cell_chat, measure = "weight")
```

In general, we can see stronger interactions between cells of the same time (diagonal in the heatmap).
This may not be surprising, as similar cells are located near each other in the tissue and thus are more likely to interact with each other.

There are also some inferred interactions across cell types.
For example, Lamp5 → Vip.
Again, these appear as neighbouring cell types in our tissue spots, so it may follow that we see some interactions between them.

```{r}
#| label: spatial-plot
#| eval: false
#| code-fold: true
#| code-summary: "Code for highlighting spatial locations of Lamp5 and Vip spots"
#| purl: false

# Highlight Lamp5 and Vip in side-by-side spatial plots
p1 <- SpatialDimPlot(
  visium,
  cells.highlight = WhichCells(visium, expression = first_type == "Lamp5"),
  image.alpha = 0.5
) +
  ggtitle("Lamp5 spot locations") +
  theme(legend.position = "none")

p2 <- SpatialDimPlot(
  visium,
  cells.highlight = WhichCells(visium, expression = first_type == "Vip"),
  image.alpha = 0.5
) +
  ggtitle("Vip spot locations") +
  theme(legend.position = "none")

p1 + p2
```

We can also visualise the ligand-receptor interactions from one cell type to others using a bubble plot.
For example, we can look at the interactions from Lamp5 to Vip and vice versa.

```{r}
#| label: visualise-bubble
#| fig-height: 12

# Bubble plot of ligand-receptor interactions from Lamp5 to Vip and vice versa
netVisual_bubble(
  cell_chat,
  sources.use = c("Lamp5", "Vip"),
  targets.use = c("Lamp5", "Vip")
)
```

## Pathway-level Visualisation

We also can focus our analysis at the level of pathways, which we have already computed the communication probabilities for.

For example, let's focus on two specific pathways of interest: "COLLAGEN" and "LAMININ".

- Laminin is a key component of the extracellular matrix and plays a crucial role in cell adhesion, differentiation, migration, and signaling.
  It is particularly important in the context of neural development and function, making it a relevant pathway to investigate in brain tissue.
- Collagen is another major component of the extracellular matrix and is involved in providing structural support to tissues.
  It also plays a role in cell adhesion and signaling, making it another important pathway to consider in the context of cell-cell interactions.

We can get a barplot of the contribution of each ligand-receptor interaction pair to these pathways using the `netAnalysis_contribution()` function.

```{r}
#| label: visualise-pathway-contribution

# barplot of the contribution of each ligand-receptor interaction pair to the selected pathways
p1 <- netAnalysis_contribution(cell_chat, signaling = "COLLAGEN") +
  ggtitle("COLLAGEN pathway")

p2 <- netAnalysis_contribution(cell_chat, signaling = "LAMININ") +
  ggtitle("LAMININ pathway")

p1 +
  p2 +
  plot_annotation(title = "Contribution of ligand-receptor interaction pairs")
```

We can compute centrality scores for selected pathways, which indicate the importance of each cell type in the communication network for those pathways.

```{r}
#| label: visualise-pathway-centrality
#| fig-height: 4

# compute centrality scores for the selected pathways
cell_chat <- netAnalysis_computeCentrality(cell_chat)

# visualise the centrality scores for a selected pathway
netAnalysis_signalingRole_network(
  cell_chat,
  signaling = "LAMININ",
  width = 16,
  height = 8
)
```

This focuses on the importance of each cell type as sender, receiver, mediator and influencer in the network.
For example, we can see that, for the LAMININ pathway, the L5 IT cell type is the most important receiver, while VLMC is the most important sender.

We can get a broader picture of each cell type's role in the overall communication network by looking at the incoming vs outgoing signalling strength for all signalling roles in the network per cell type.

```{r}
#| label: visualise-signaling-roles

# All pathways
netAnalysis_signalingRole_scatter(cell_chat)

# Only the LAMININ pathway
netAnalysis_signalingRole_scatter(cell_chat, signaling = c("LAMININ"))
```

In general, there is a correlation between the strenght of outgoing and incoming signalling, which makes sense as cell types that are more active in sending signals are also likely to be more active in receiving signals.

However, for the LAMININ pathway example specifically, we can see that the L5 IT cell type is more active in receiving signals than sending signals, while VLMC is more active in sending signals than receiving signals, which is consistent with the centrality scores we observed above.

## Spatial Visualisation of Pathways

Finally, we can visualise specific signalling pathways of interest in spatial context using the `netVisual_aggregate()` function with the `layout = "spatial"` option.

```{r}
#| label: visualise-spatial-pathway

netVisual_aggregate(
  cell_chat,
  signaling = "LAMININ",
  layout = "spatial",
  edge.width.max = 2,
  vertex.size.max = 1,
  alpha.image = 0.2,
  vertex.label.cex = 3.5
)
```

The spatial plots visualize the selected signaling pathways in the context of the tissue.
The nodes represent different cell types, and the edges represent the interactions between them.
The layout is based on the spatial coordinates of the spots, allowing us to see how cell-cell interactions are distributed across the tissue.

There is a plethora of other visualization and investigation options available in the CellChat package.
You can explore these options in the CellChat documentation to find the best way to visualize your specific data and research questions.

## Caveats and Limitations

It is important to remember that ligand-receptor interaction networks derived from single-cell/spatial transcriptomics represent **potential communication flows, not direct measurements of signalling activity**.
Most approaches, including CellChat, use gene expression measurements together with a curated database of known ligand-receptor interactions to identify signalling relationships that are consistent with the observed data.

One key limitation is that **mRNA expression is only a proxy for the abundance and activity of signalling proteins**.
Expression of a ligand does not mean that the protein is produced and secreted, while expression of a receptor does not guarantee that the signalling pathway is active.
Many signalling processes are regulated at the post-transcriptional level (e.g. protein modifications, receptor localisation, etc.), which is not captured by transcriptomics data.

The quality of CCC results also depends heavily on the input data, in particular the **cell annotations**.
Cell type annotations are particularly important because communication is typically inferred between cell populations.
Incorrect annotations can obscure biologically meaningful interactions, while over-clustering may introduce artificial interactions between closely related cell states.
Finally, **the choice of ligand-receptor database** can have a substantial impact on the interactions identified, as different databases contain different interaction sets and levels of curation.

For spatial transcriptomics data, physical proximity can provide additional evidence supporting a potential interaction.
However, spatial proximity alone does not demonstrate communication, and communication can sometimes occur over longer distances.
Furthermore, lower-resolution technologies may contain mixtures of multiple cell types within a single spot, making it more difficult to accurately assign signalling interactions to individual cells.

Finally, **CCC results should be used as hypothesis-generating** rather than definitive evidence of signalling.
Interactions identified this way should be complemented by biological knowledge and validated in the lab, if possible.

## Summary

::: {.callout-tip}
#### Key Points

- Cell-cell interactions are crucial for understanding tissue microenvironments and biological processes.
- The `CellChat` package provides a comprehensive framework for analyzing and visualizing cell-cell communication networks.
- Preparing the Seurat object and identifying overexpressed genes and interactions are essential steps in the analysis.
- For spatial transcriptomics data, incorporating spatial coordinates can enhance the analysis of cell-cell interactions.
- Visualization techniques such as circle plots, heatmaps, and bar plots help interpret the results of cell-cell interaction analysis.
:::
