# Deconvolution

```{r}
#| label: setup
#| include: false
#| purl: false

knitr::opts_chunk$set(
  echo = TRUE,
  message = FALSE,
  warning = FALSE,
  cache.lazy = FALSE
)

# set the working directory to the course_files folder
# make sure to run course_files/download_course_files.sh
knitr::opts_knit$set(root.dir = here::here("course_files"))
```

::: {.callout-tip}
### Learning Objectives

- Explain why deconvolution is needed for lower-resolution spatial transcriptomics data and how spot resolution affects interpretation.
- Prepare compatible reference and query inputs for `spacexr` from Seurat objects.
- Run `create.RCTD()` and `run.RCTD()` with appropriate `doublet_mode` settings for the biological and technical context.
- Extract RCTD classifications and weights and add them to Seurat metadata for downstream analysis.
- Interpret and compare deconvolution outputs using spatial plots, thresholds, and heatmaps in the context of known tissue biology.
:::

## Setup

We continue working with the sagittal mouse brain dataset from previous chapters.

::: {.callout-note collapse="true"}
#### Click to expand

Start by loading the required libraries and the Seurat object.

```{r}
#| label: load-visium-data

# Load libraries
library(Seurat) # single-cell and spatial analysis toolkit
library(sparseMatrixStats)
library(paletteer) # colour palettes
library(ggplot2) # plotting
library(dplyr) # data manipulation
library(patchwork) # combining plots

# Load the Seurat object from the previous chapter
visium <- readRDS("precomputed/mouse_brain_banksy.rds")
```

As a reminder, this object was created by:

- Importing the raw data using `Load10X_Spatial()`, followed by QC and filtering to remove low-quality spots.
- Data normalisation using `SCTransform()`, which models technical variation and applies variance stabilisation.
  Stored in the "SCT" assay.
- Applying dimensionality reduction using PCA, UMAP and t-SNE.
- Clustering the data using graph-based clustering.
  The clusters we use in this chapter were generated using the Leiden algorithm at a resolution of 0.8 and were set the the default cell identities (`Idents(visium)`).
- Identifying spatially variable features using Moran's I and markvariogram methods.
- Running the BANKSY algorithm to identify spatial domains in the data, which were stored in the `banksy_clusters` column of the metadata.
:::

## Overview

Deconvolution estimates the cellular composition of mixed gene expression profiles.
In spatial transcriptomics, this means estimating which cell types contribute to each pixel or spot.
This is especially important for lower-resolution technologies, where each spot can capture multiple cells (for example Visium V1/V2 spots with 55 μm diameter).
When those cells belong to different types, the measured expression is a mixture, which can affect downstream analyses such as clustering and tissue architecture analysis.

Several tools are available for spatial deconvolution, reviewed in [Gaspard-Boulinc et al. 2022](https://doi.org/10.1038/s41467-022-29913-1).
Here we use Robust Cell Type Decomposition (RCTD) from [Cable et al. 2022](https://doi.org/10.1038/s41587-021-00830-w), which was one of the top-performing methods in that benchmark.

Briefly, RCTD requires:

- an annotated reference dataset (for example single-cell RNA-seq)
- a query spatial transcriptomics dataset to deconvolute

It then uses a probabilistic model to infer cell type mixtures in each spot.

RCTD is implemented in the `spacexr` package.
We also load `pheatmap` for heatmap visualisations later in the chapter.

```{r}
#| label: load-libraries

# Load necessary libraries
library(spacexr)
library(pheatmap)
```

::: {.callout-warning collapse="true"}
#### `spacexr` package versions

Confusingly, there are two version of this package:

- The original package available from GitHub ([dmcable/spacexr](https://github.com/dmcable/spacexr)) has its own object structure and implements other methods not covered here.
- A [re-implementation in Bioconductor](https://www.bioconductor.org/packages/release/bioc/html/spacexr.html) using its standard `SpatialExperiment`/`SingleCellExperiment` objects.
  This package only implements the RCTD analysis of the original package.

While both provide similar functionality for RCTD analysis, the syntax is slightly different.
We are using the original GitHub version here, so be aware that adaptations may be needed if you use the Bioconductor version instead.
:::

## Reference Dataset

The first step for deconvolution is choosing a suitable reference dataset.
That choice depends on tissue, organism and data availability, and well-annotated references may be lacking for non-model organisms.

Here we use a preprocessed single-cell RNA-seq dataset from the [Allen Brain Cell (ABC) Atlas](https://brain-map.org/bkp/explore/abc-atlas).
Cells are annotated hierarchically as class → subclass → cluster, from broad categories to specific cell populations.

- **Class**: broad categories such as glutamatergic neurons, GABAergic neurons, astrocytes, and oligodendrocytes.
- **Subclass**: groups within a class that share similar transcriptional profiles (for example Vip, Sst, and Lamp5 within GABAergic neurons).
- **Cluster**: the most specific level, often named with distinguishing marker genes (for example *Vip Arhgap36 Hmcn1*).

We start by loading the reference object and look at its structure.
Because this is a precomputed object, we read it from RDS and run `UpdateSeuratObject()` for compatibility with current Seurat versions.

```{r}
#| label: load-reference

# Load the reference single-cell dataset
ref <- readRDS("precomputed/mouse_brain_reference_for_RCTD.rds")
ref <- UpdateSeuratObject(ref)

# Assays present in object
Assays(ref)
```

This object includes an "RNA" assay with raw counts.
It also includes an "SCT" assay containing sctransform-normalised data, set as the default assay.

Metadata includes the three annotation levels: `class`, `subclass` and `cluster`.

```{r}
#| label: reference-annotations

# Check the annotation levels in the reference dataset
ref[[]] |>
  select(class, subclass, cluster) |>
  head()

# Count how many of each annotation level we have
length(unique(ref$class))
length(unique(ref$subclass))
length(unique(ref$cluster))
```

We can visualise these annotations on a UMAP to see how they map onto cell populations.

```{r}
#| label: reference-umaps
#| code-fold: true
#| code-summary: "Click to expand visualisation code"

# Visualise UMAP and Spatial plots of the reference dataset with different annotation levels
umap_class <- DimPlot(
  ref,
  reduction = "umap",
  group.by = "class",
  label = TRUE
)
umap_subclass <- DimPlot(
  ref,
  reduction = "umap",
  group.by = "subclass",
  label = TRUE
)
umap_cluster <- DimPlot(
  ref,
  reduction = "umap",
  group.by = "cluster",
  label = FALSE
)

# Assemble the UMAPs together with common settings
(umap_class + umap_subclass + umap_cluster) &
  coord_equal() &
  theme_void() &
  theme(legend.position = "none")
```

This can help us choose the annotation level for deconvolution.
In this chapter we use the subclass level (`r length(unique(ref$subclass))` categories), which gives a practical balance between granularity and interpretability.

Before running RCTD, check how many reference cells are available per subclass.
If you use a different reference dataset, ensure each cell type has enough cells (at least 25) for a reliable estimation by RCTD.

```{r}
#| label: reference-cell-counts
#| results: "hide"

# Count number of reference cells in each subclass
table(ref$subclass)
```

## Preparing RCTD Data

RCTD requires both reference and query data in its own object formats.

### Reference Object

The reference object needs:

- a count matrix
- cell type labels
- total UMI counts per cell

We extract these from the Seurat object and construct a `Reference` object.

```{r}
#| label: prepare-reference
#| cache: true

# raw counts from the RNA assay
ref_counts <- ref[["RNA"]]$counts

# our cluster identities will be the subclass annotation
ref_cluster <- as.factor(ref$subclass)
names(ref_cluster) <- colnames(ref)

# the total number of UMIs for each cell
ref_umis <- ref$nCount_RNA
names(ref_umis) <- colnames(ref)

# Create the Reference object for RCTD
reference <- Reference(ref_counts, ref_cluster, ref_umis)

# Remove intermediate objects to free memory
rm(ref, ref_counts, ref_cluster, ref_umis)
```

### Query Object

Similarly, we prepare the spatial dataset to deconvolute.
RCTD needs:

- raw counts from the "Spatial" assay
- spot coordinates
- total UMI counts per spot

```{r}
#| label: prepare-query
#| cache: true

# raw counts
query_counts <- visium[["Spatial"]]$counts

# spot coordinates
query_coords <- GetTissueCoordinates(visium)[, c("x", "y")]

# total UMIs for each spot
query_umis <- colSums(query_counts)

# Construct the query object for RCTD
query <- SpatialRNA(query_coords, query_counts, query_umis)

# remove intermediate objects to free memory
rm(query_counts, query_coords, query_umis)
```

## Running RCTD

With reference and query objects prepared, we can run deconvolution in two steps:

- `create.RCTD()`: builds the RCTD object and prepares model inputs.
  This includes estimating cell type profiles from the reference, selecting informative genes (differentially expressed between classes), and estimating a normalisation factor to account for platform differences in the reference (e.g. single-cell versus single-nucleus).
- `run.RCTD()`: fits the model and estimates cell type proportions for each spot.

Run the first step to create the RCTD object:

```{r}
#| label: create-rctd
#| results: "hide"

# Create RCTD object on 8 cores
rctd <- create.RCTD(query, reference, max_cores = 8)

# Remove intermediate objects to free memory
rm(query, reference)
```

We set **`max_cores = 8`** to reduce runtime, which can be substantial for larger datasets.

RCTD supports three deconvolution modes through `doublet_mode`:

- `"doublet"`: allows 1-2 cell types per spot.
  This is often suitable for higher-resolution spatial technologies.
- `"multi"`: similar to doublet mode but allows more than two cell types, up to `MAX_MULTI_TYPES` in `create.RCTD()` (default 4).
- `"full"`: places no explicit upper limit on the number of assigned cell types.

Constrained modes ("doublet" and "multi") can reduce overfitting, where many very small proportions are assigned to each spot.
The best choice depends on tissue complexity and spatial resolution of the technology.

We start with `"doublet"` mode and later compare results with `"full"` mode.
RCTD is computationally demanding, so we provide a precomputed object.
You can expand the code block below to see the command used.

```{r}
#| label: run-rctd-doublet
#| cache: true
#| code-fold: true
#| code-summary: "Click for `run.RCTD()` command - takes a long time to run!"
#| results: "hide"

# This will run about ~12-15 minutes on 8 cores
rctd <- run.RCTD(rctd, doublet_mode = "doublet")
```

```{r}
#| label: save-rctd-doublet
#| include: false
#| purl: false

if (!file.exists("precomputed/mouse_brain_rctd_doublet.rds")) {
  saveRDS(rctd, "precomputed/mouse_brain_rctd_doublet.rds")
}
```

```{r}
#| label: read-precomputed-rctd-doublet

# Read precomputed rctd object in doublet mode
rctd <- readRDS("precomputed/mouse_brain_rctd_doublet.rds")
```

We can now add the RCTD results to the Seurat object and for further analysis.
Because we ran in doublet mode, we expect output describing major and secondary cell types per spot.

## Extracting deconvolution results

The RCTD object stores results in several components.
We extract the relevant parts into a data frame and then add them to the Seurat metadata.

We will extract:

- spot-level classification
- cell type weights (estimated proportions)

These values are in `rctd@results`:

```{r}
#| label: rctd-doublet-results-df

# Main results data frame - retain only some columns of interest
doublet_results <- rctd@results$results_df |>
  select(spot_class, first_type, second_type)

# Add the associated with each type
doublet_results <- cbind(doublet_results, rctd@results$weights_doublet)

# Ensure column names are unique
names(doublet_results) <- c(
  "spot_class",
  "first_type",
  "second_type",
  "first_weight",
  "second_weight"
)

# Inspect the table
head(doublet_results)
```

Column meanings:

- `spot_class`: spot classification (`"singlet"`, `"doublet_uncertain"`, `"doublet_certain"`, `"reject"`).
- `first_type` and `second_type`: the top two cell type assignments (using reference `subclass` labels).
- `first_weight` and `second_weight`: estimated proportions for those two assigned types.

Start by counting spots in each class:

```{r}
#| label: count-doublet-class

# Count spot classes
doublet_results |>
  count(spot_class)
```

We have no spots classified as "singlet", which is expected given the spot size of Visium v1 and the complexity of the tissue.
A high `"doublet_certain"` count indicates many spots are confidently modelled as mixtures of at least two types.
`"doublet_uncertain"` and `"reject"` spots are useful to look at as well, because they can indicate ambiguous expression profiles or reference mismatches.

We can add these classifications to our Seurat metadata and visualise their spatial distribution:

```{r}
#| label: add-doublet-to-seurat

# Add results back into visium object
visium <- AddMetaData(visium, metadata = doublet_results)

# Check where each spot is classified as a doublet
# We use a custom colour scheme to better distinguish the different classes
SpatialDimPlot(visium, group.by = "spot_class") +
  scale_fill_manual(
    values = c(
      "reject" = "red",
      "doublet_certain" = "steelblue",
      "doublet_uncertain" = "orange"
    )
  ) +
  ggtitle("Spot classification by RCTD")
```

We can also compare RCTD type assignments with the Leiden clusters identified earlier:

```{r}
#| label: visualise-doublet-types
#| code-fold: true
#| code-summary: "Click for visualisation code"

# Spatial plot for first type
first_type_plot <- SpatialDimPlot(
  visium,
  group.by = "first_type"
) +
  ggtitle("First cell type identified by RCTD")

# Spatial plot for second type
second_type_plot <- SpatialDimPlot(
  visium,
  group.by = "second_type"
) +
  ggtitle("Second cell type identified by RCTD")

# Leiden clusters
leiden_cluster_plot <- SpatialDimPlot(
  visium,
  group.by = "Leiden_08",
  label = TRUE,
  label.size = 4
) +
  theme(legend.position = "none") +
  ggtitle("Clusters (Leiden)")

# Assemble the plots together
(first_type_plot + second_type_plot) / leiden_cluster_plot
```

## Full cell type analysis

So far we have focused on doublet mode, which reports the top one or two cell types per spot.
This can be useful when spots are likely to contain only a small number of cell types.

For Visium v1, each 55 μm spot often contains multiple cells from several types.
In this case, `"full"` mode may provide a more informative view of cell type composition in our spots.

Because this run is slow, we provide a precomputed file.
The code below shows how it was generated.

```{r}
#| label: run-rctd-multi
#| cache: true
#| code-fold: true
#| code-summary: "Click for `run.RCTD()` command - takes a long time to run!"
#| results: "hide"

# Run RCTD in full mode
# Takes a long time to run! Read the precomputed object below instead
rctd_full <- run.RCTD(rctd, doublet_mode = "full")
```

```{r}
#| label: save-rctd-multi
#| include: false
#| purl: false

if (!file.exists("precomputed/mouse_brain_rctd_full.rds")) {
  saveRDS(rctd_full, "precomputed/mouse_brain_rctd_full.rds")
}
```

```{r}
#| label: read-precomputed-rctd-multi
#| eval: false

# Read precomputed rctd object in doublet mode
rctd_full <- readRDS("precomputed/mouse_brain_rctd_full.rds")
```

::: {.callout-note collapse="true"}
#### Should I always run RCTD in "full" mode?

Although it might sound like a good idea to always run the model in "full" mode, it is important to remember that the model may overfit the data, so it's not a bad idea to restrict to doublets when working with high-resolution data.

In our case, as we are working with a low-resolution technology, it makes sense to attempt running the analysis in full mode.
:::

### Extract and visualise cell weights

In full mode, RCTD returns a weight matrix in `rctd@results$weights`.
Rows correspond to spots, columns correspond to reference cell types, and each value is the estimated proportion of a cell type in a spot.

This can be used for two common tasks:

- broad exploration of all cell types
- analysis of specific types of interest

The next code block extracts and reshapes these weights for visualisation with ggplot2.
We plot all subclasses to get a global view of composition patterns.

```{r}
#| label: visualise-weights

# Extract weights
rctd_weights <- rctd_full@results$weights

# Add coordinates
rctd_weights <- cbind(rctd_weights, GetTissueCoordinates(visium)[, c("x", "y")])

# Convert to tibble and pivot to long format
rctd_weights <- rctd_weights |>
  as_tibble(rownames = "spot_barcode") |>
  tidyr::pivot_longer(
    -c(spot_barcode, x, y),
    names_to = "subclass",
    values_to = "weight"
  )

# Visualise weights for each subclass
# Note we're capping the colour scale to 0.5 to better visualise differences
# We also use negative y values to match the orientation of the spatial plot
rctd_weights |>
  ggplot(aes(x, -y, colour = weight)) +
  geom_point(size = 0.1) +
  facet_wrap(~subclass) +
  theme_void() +
  scale_colour_viridis_c(limits = c(0, 0.5))
```

As always, you should **interpret these results grounded in biological knowledge** of the tissue and cell types you are looking at.
You can focus on whether high-weight regions align with expected anatomical locations.
Unexpected patterns can reflect genuine biology, reference limitations, or model overfitting.

For example, consider cell type "L2_3_IT" (layer 2/3 intratelencephalic neurons; see [ABC Atlas entry](https://knowledge.brain-map.org/celltypes/CCN202002013/CS202002013_183)):

```{r}
#| label: highlight-rctd-cells
#| code-fold: true

# Highlight a particular cell type
SpatialPlot(
  visium,
  cells.highlight = WhichCells(visium, expression = first_type == "L2_3_IT")
)
```

These spots appear in superficial cortical layers, which is biologically plausible.
However, we also see a strong signal in inner tissues, which may be less expected.

Here are a couple of reasons why you might see a signal in less expected regions:

- Model overfitting assigns low-confidence contributions, when in fact that cell type is absent.
- The reference lacks true matching cell states, so the model assigns the closest available type.

For this reason, treat deconvolution as one source of evidence and cross-check with marker genes, spatial context and tissue knowledge.

### Add cell weights to Seurat

As with doublet-mode results, we can add full-mode weights to Seurat metadata:

```{r}
#| label: add-rctd-full-to-metadata

# Add cell type weights to metadata
visium <- AddMetaData(visium, rctd_full@results$weights)

# Check this has been added as new columns to metadata
colnames(visium[[]])
```

You now have one metadata column per cell type, containing RCTD weight estimates.
These columns can be used directly in spatial plots or dimensionality-reduction plots.

### Identify most abundant type

We can also assign each spot its most abundant inferred cell type.
Using the long-format weights table, we select the highest-weight subclass per spot and add it back to metadata.

```{r}
#| label: add-max-type-to-metadata

max_cell_type <- rctd_weights |>
  # for each barcode
  group_by(spot_barcode) |>
  # get the cell type with the highest weight
  slice_max(weight) |>
  ungroup()

# add to metadata - make sure to match the order of the barcodes
visium$rctd_max_type <- max_cell_type$subclass[match(
  colnames(visium),
  max_cell_type$spot_barcode
)]
visium$rctd_max_weight <- max_cell_type$weight[match(
  colnames(visium),
  max_cell_type$spot_barcode
)]

# Visualise the most abundant cell type in each spot
SpatialDimPlot(visium, group.by = "rctd_max_type") +
  ggtitle("Most abundant cell type in each spot")
```

You can also apply a confidence threshold, for example only assigning a label if the maximum weight is above 0.5:

```{r}
#| label: add-threshold-to-max-type

# add metadata column indicating whether the max weight is greater than 0.5
visium$rctd_max_type_gt_05 <- ifelse(
  visium$rctd_max_weight > 0.5,
  visium$rctd_max_type,
  NA
)

# Visualise the most abundant cell type with weight > 0.5
SpatialDimPlot(visium, group.by = "rctd_max_type_gt_05") +
  ggtitle("Most abundant cell type in each spot (weight > 0.5)")
```

If very few spots pass this threshold, that suggests most spots are mixed, which is expected for Visium v1.
It can also indicate that full mode is distributing small weights across many candidate types.

You can explore these thresholds to explore the deconvolution results in different ways, and to identify the most likely cell type in each spot based on the weights assigned by RCTD.

### Weight Heatmaps

We can also inspect cell type proportions as a heatmap.
One useful strategy is ordering spots by previous clustering labels.
Here we order spots by the `Leiden_08` metadata column from the previous chapter.

```{r}
#| label: heatmap-weights

# prepare heatmap matrix from weights matrix
# order the rows (spots) based on Leiden clusters
heatmap_data <- rctd_full@results$weights[order(visium$Leiden_08), ]

# Transpose the matrix for easier heatmap interpreation (cell types as rows, spots as columns)
heatmap_data <- t(heatmap_data)

# Visualise as heatmap adding leiden cluster annotation
pheatmap(
  heatmap_data,
  cluster_rows = FALSE,
  cluster_cols = FALSE,
  show_colnames = FALSE,
  annotation_col = data.frame(Leiden_08 = sort(visium$Leiden_08)),
  main = "Cell Type Proportions"
)
```

We can see some correlation between the Leiden clusters and the cell type proportions, which is expected as both are based on gene expression profiles.
But this is not a perfect alignment, we can see many scattered spots with high proportions of certain cell types that do not correlate with these clusters.
We might expect this result as each spot likely contains multiple cell types, and the clustering is based on the overall gene expression profile of the spot, which can be confused by the presence of multiple cell types.

Alternatively, you can cluster the heatmap columns to group spots by composition similarity:

```{r}
#| label: heatmap-cluster-columns

# Heatmap with clustered spots
pheatmap(
  heatmap_data,
  cluster_rows = FALSE,
  cluster_cols = TRUE,
  show_colnames = FALSE,
  annotation_col = data.frame(Leiden_08 = sort(visium$Leiden_08)),
  main = "Cell Type Proportions"
)
```

This view shows spot groups that do not perfectly match expression-defined clusters.

For example, there is a small cluster of spots with high "L6b" proportions:

```{r}
#| label: spatial-L6b
#| code-fold: true
#| fig-height: 4

# Spatial plot for L6b weight
spatial_l6b <- SpatialFeaturePlot(visium, features = "L6b") +
  ggtitle("L6b cell type")

# UMAP with these cells highlighted
umap_l6b <- DimPlot(
  visium,
  reduction = "umap",
  cells.highlight = WhichCells(visium, expression = rctd_max_type == "L6b")
) +
  theme(legend.position = "none") +
  coord_equal() +
  theme_void()

spatial_l6b + umap_l6b
```

These spots form a narrow spatial band, often close to a single line of spots.
That pattern is consistent with layer 6b neurons occupying a thin cortical layer.

This example also shows why analysis resolution matters.
In the previous [spatial clustering chapter](09-SpatialClustering.qmd), BANKSY used `k_geom = 18`, averaging spatial features across 18 neighbouring spots (two rings of spots around each spot).
For very narrow structures, this averaging can dilute local signals and make small sub-domains harder to identify.

This again highlights the importance of **biological knowledge** of your tissue and considering the **resolution of the technology** used.

## Exporting Seurat Object

Optionally, save the Seurat object with deconvolution results for later use:

```{r}
#| label: save-seurat-object
#| eval: false
#| purl: false

# Save Seurat object
saveRDS(visium, "results/mouse_brain_visium_deconvolution.rds")
```

```{r}
#| label: save-participant-data
#| include: false
#| purl: false

# Save participant data
if (!file.exists("precomputed/mouse_brain_visium_deconvolution.rds")) {
  saveRDS(visium, "precomputed/mouse_brain_visium_deconvolution.rds")
}
```

## Summary

::: {.callout-tip}
### Key Points

- Deconvolution helps identify cell type mixtures in spatial spots where multiple cells contribute to one expression profile.
  - Spot size and tissue complexity determine how strongly mixed each spot is likely to be.
  - Mode choice (`doublet` versus `full`) depends on the expected number of contributing cell types.

- `spacexr` requires reference and query objects for model fitting.
  - `Reference()` uses raw counts, cell labels, and per-cell UMIs from the single-cell reference.
  - `SpatialRNA()` uses spot coordinates, raw counts, and per-spot UMIs from the spatial dataset.

- RCTD runs in two stages: object construction with `create.RCTD()` and deconvolution with `run.RCTD()`.

- Doublet-mode outputs gives spot classes and top cell type assignments that can be integrated into Seurat metadata.
  - `spot_class` distinguishes confident doublets, uncertain doublets, and rejected spots.
  - `first_type`/`second_type` and their weights provide the two inferred cell types and their frequencies.

- Full-mode weights can be used for a more exhaustive analysis of cell types.
  - Heatmaps and spatial plots can be used to compare composition patterns with previous clustering results.
  - Results should always be compared to known tissue biology and interpreted with caution.
:::
